---
title: "GLS supervised calibration analyses"
author: "Diego Vicente-Sastre"
format: html
editor: visual
date-modified: today
#bibliography: references.bib
execute: 
  echo: true
  warning: false
  message: false
---

## 1. Introduction: why calibrate GLS?

Twilight events are transitions of light between day and night, and *vice versa*. To establish them, it is set a predetermined light intensity threshold that divides daytime from night, so sunrises and sunsets can be temporarily noted (threshold-based method). The threshold must consist in the minimum possible value that is consistently above the noise in the light intensity captured at night (Lisovski et al., 2019).

GLS may vary in the way they measure and capture light due to environmental reasons or accuracy errors. Thus, it is first necessary to carry out a calibration of these devices to relate the expected light levels to those observed and be able to extract the final positions with minimal error. We employ the rooftop calibration method (Lisovski and Hahn, 2012). Each GLS has at least two calibrations, one pre-deployment and one post-recovery, in stationary locations with a minimum of light pollution. These calibrations should last at least seven days each, with the devices placed over a flat elevated surface, considering there cannot be any element that shadows (e.g., from buildings or rock walls), and in two different geographical locations, if possible.

The sunrise and sunset times captured by the GLS are associated with those of the area where the calibration was carried out. This information must be manually reviewed to correct or remove misplaced twilight events due to false light peaks in non-transition hours. At the same time, points on the map move with the twilight correction until they are all adjusted to the correct location. These calibrations involve subjective assessment, but Bråthen et al. (2021) showed that it could be performed consistently and with a high repeatability.

## 2. Load packages and functions

Some packages require to be installed remotely from other repositories outside of CRAN. Following lines need to be run only the first time.

```{r install_pkgs, include=TRUE, eval=FALSE}
# Both packages allow us to download packages from external repositories as GitHub
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")

pacman::p_load(devtools, remotes, install = TRUE)

pacman::p_load_gh(
  "SWotherspoon/SGAT",
  "SWotherspoon/BAStag",
  "bmcclintock/momentuHMM",
  "SLisovski/TwGeos",
  "SLisovski/GeoLight"
)
```

We load other required packages.

```{r loading_func}
pacman::p_load(
  'devtools', 'remotes',        # installing external packages
  'readxl',                     # opening Excels and csv
  'tidyverse',                  # managing data
  'lubridate',                  # dealing with dates
  'evaluate',                   # check errors
  'maps', 'rworldmap',          # creating maps for GeoLight
  'GeoLight', 'BAStag', 'TwGeos' # geolocators
)

# Detach purrr to avoid conflicts with base map()
if ("purrr" %in% .packages()) {
  detach("package:purrr", unload = TRUE)
}
```

Lastly, we load own functions, specifically designed for working with GLS data.

```{r extra_func}
source("functions/01_FUNCTIONS_GLS_analysis.R")
```

## 3. Read metadata and define thresholds

```{r load_data}
mdata      <- read_csv('input/GLS.files.csv',        col_names = TRUE)
info_file  <- read_csv('input/GLS.info.csv',         col_names = TRUE)
calib.info <- read_csv('input/Info.calibrations.csv')
```

**Metadata file** contains the files available for each GLS (i.e., light, wet-dry and SST data) and the format of each file, which depends on the producer.

**Calibrations information file** contains the specific information for each calibration period of each GLS. Each GLS can be calibrated more than once and in different places. Here are added the coordinates of each calibration locality.

```{r set_params}
# Light threshold: value above which it is daytime (typically 1.5-2)
threshold <- 2

# Map extent — used as visual aid during calibration
xlim <- c(-20, 0)
ylim <- c(15, 40)
map  <- rworldmap::getMap(resolution = "coarse")

# Method for thresholdCalibration: "SGD" or "IRLS"
method <- "SGD"
```

**Threshold** defines the light level above which it is day and below it is night. Normally it is between 1.5 and 2. The **map** will be a visual aid, but not the main reference for decisions. The **method** controls the fitting algorithm used inside `thresholdCalibration`.

## 4. Extraction of calibration parameters

For each GLS, up to three calibration periods are processed:

- **Steps 1–2**: pre-deployment calibrations (suffix `_PD1`, `_PD2`)
- **Step 3**: post-recovery calibration (suffix `_PR3`)

We extract the following parameters via `thresholdCalibration` from the `TwGeos` package:

- **Zenith angles**: zero-deviation (`zenith0`) and median-deviation (`zenith`)
- **Error distribution parameters**: shape (`alpha.mean`) and scale (`alpha.sd`)

### 4.1. Read GLS information

We get all the **essential information** for a single GLS (here `i = 1`). In a full analysis this block runs inside an outer loop over all rows of `mdata`.

```{r GLS_info}
i <- 1
gls.dir <- paste0(getwd(), '/input/GLS/')

info       <- mdata[i, ]
lightfile  <- if_else(!is.na(info$File.adjlux), info$File.adjlux, info$File.lux)
lightfile_c <- paste0(gls.dir, lightfile)

geoID    <- info$geo_alfanum
Producer <- info$Producer   # e.g. "Migrate Technology", "Biotrack"
col.name <- info$Colony
sp       <- info$Sp
year_rec <- info$Year_rec

cat('Processing geo (', i, ')', geoID, 'for year', year_rec, '\n')

# Calibration info for this GLS
calib <- calib.info %>% filter(geo_alfanum == geoID)

if (nrow(calib) == 0) stop("No calibration info for geo ", geoID, " — check Info.calibrations.csv")
```

We can also extract additional metadata useful for saving outputs or working across colonies and species.

```{r GLS_extra_info}
start_date <- as.Date(info$start_date, tz = 'GMT')
end_date   <- as.Date(info$end_date,   tz = 'GMT')
```

### 4.2. Read light data

```{r read_light}
lu <- read_lightGLS(
  producer = Producer,
  ID       = geoID,
  file     = lightfile_c
  # start_date = start_date  # uncomment to subset to deployment period
  # end_date   = end_date
)
```

### 4.3. Calibration parameters

For each of the (up to) three calibration periods we run the interactive `calib_period` function — which displays the light image and lets the user review twilight events — and then extract SGAT parameters via `thresholdCalibration`.

> **⚠️ Interactive step — run chunk by chunk from RStudio, do not use *Render*.** `calib_period` opens a plot window that requires manual interaction (clicking to accept or correct twilight events). Make sure all previous chunks have been run first so that `lu`, `calib`, `geoID`, etc. are available in your environment.

```{r calib_init}
# Initialise accumulator lists before the interactive loop
# <<- assigns to the global environment so results survive across chunk runs
calib.list_SGAT <<- vector('list', nrow(mdata))
twl_error.list  <<- list()
```

```{r main_loop}
# ── Run this chunk interactively from RStudio (not via Render) ──────────────
# calib_period() opens an interactive plot window. After reviewing each light
# image, close the window to advance to the next calibration step.

for (step in 1:3) {

  calib_start <- as.POSIXct(calib[[paste0("calib_start", step)]], tz = "GMT")
  calib_end   <- as.POSIXct(calib[[paste0("calib_end",   step)]] + 1, tz = "GMT")
  n_days <- round(as.numeric(difftime(calib_end, calib_start, units = 'days')))
  loc    <- calib[[paste0("loc_calib", step)]]
  lon    <- calib[[paste0("Lon", step)]]
  lat    <- calib[[paste0("Lat", step)]]

  # Column suffix consistent with downstream scripts: PD1, PD2, PR3
  step_label <- if (step %in% c(1, 2)) paste0("PD", step) else "PR3"
  step_name  <- if (step %in% c(1, 2)) "predep" else "postrec"

  calib_SGAT <- c(NA_real_, NA_real_, NA_real_, NA_real_)

  # length() == 0 when column has no value; is.na() == TRUE when value is NA
  if (length(loc) > 0 && length(calib_end) > 0 &&
      !is.na(loc)  && !is.na(calib_end)) {

    # Open an external graphics window: the RStudio plot pane does not support
    # mouse/keyboard event handling required by preprocessLight() inside calib_period()
    if (.Platform$OS.type == "windows") {
      windows()
    } else if (Sys.info()[["sysname"]] == "Darwin") {
      quartz()
    } else {
      x11()
    }

    # Interactive light image — review twilights, then close the window to continue
    twl <- calib_period(
      data      = lu,
      start     = calib_start,
      end       = calib_end,
      lon       = lon,
      lat       = lat,
      threshold = threshold
    )

    # Close the external device and return focus to RStudio
    if (dev.cur() > 1) dev.off()

    if (!is.data.frame(twl) || nrow(twl) == 0) {
      message('No valid twilights for step ', step)

    } else {

      # Save .trn file for this calibration period
      trn <- twl %>%
        mutate(
          date_time  = Twilight,
          type       = ifelse(Rise == TRUE, 'Sunrise', 'Sunset'),
          confidence = Marker
        ) %>%
        dplyr::select(date_time, type, confidence)

      if (!dir.exists('output/calibrations'))
        dir.create('output/calibrations', recursive = TRUE, showWarnings = FALSE)

      file_trn <- paste(geoID, sp, year_rec, col.name, step_name, '.trn', sep = '_')
      write.table(trn,
                  file      = paste0('output/calibrations/', file_trn),
                  sep       = ',',
                  col.names = FALSE,
                  row.names = FALSE,
                  quote     = FALSE)
      cat('Trn saved for calibration period:', step_name, '\n')

      # SGAT threshold calibration
      calib_SGAT <- try(
        thresholdCalibration(twl$Twilight, twl$Rise, lon, lat, method = method),
        silent = TRUE
      )
      if (inherits(calib_SGAT, 'try-error')) {
        message('Error in SGAT calibration at step ', step)
        calib_SGAT <- c(NA_real_, NA_real_, NA_real_, NA_real_)
      }
    }

  } else {
    message('No calibration information for step ', step)
  }

  # Build one row for this calibration step
  twl_error <- bind_cols(
    geoID      = geoID,
    Producer   = Producer,
    Year       = year_rec,
    Colony     = col.name,
    n_days     = n_days,
    zenith     = calib_SGAT[1],
    zenith0    = calib_SGAT[2],
    alpha.mean = calib_SGAT[3],
    alpha.sd   = calib_SGAT[4],
    Loc        = loc
  )
  colnames(twl_error)[5:10] <- paste0(colnames(twl_error)[5:10], '_', step_label)

  twl_error.list[[step]] <<- twl_error
}

# Collapse the 3 steps into one row per GLS and store in global list
twl_error_GLS <- bind_rows(twl_error.list) %>%
  group_by(geoID, Producer, Year, Colony) %>%
  summarise(across(everything(), ~ na.omit(.x)[1]), .groups = 'drop')

calib.list_SGAT[[i]] <<- twl_error_GLS
cat('Geo (', i, ')', geoID, 'done — move to next GLS by changing i and re-running from 4.1\n')
```

```{r join_calib, eval=FALSE}
# ── Run this chunk only after all GLS have been processed ───────────────────
calib.all_SGAT <- bind_rows(calib.list_SGAT)

str(calib.all_SGAT)
# Convert parameter columns to numeric (already numeric from loop, just enforce type)
calib.all_SGAT <- calib.all_SGAT %>%
  mutate(across(c(3, 5:8, 10:13, 15:18), as.numeric))

# Save intermediate result before outlier removal
write_csv(calib.all_SGAT, file = 'users_guide/Manual_calibrations_SGAT.csv')
cat('Raw calibration results saved.
')
```

The chunk below loads the results already saved from a previous interactive session, so the rest of the document can be rendered without re-running the interactive steps.

```{r calib_results}
calib.all_SGAT <- read_csv("users_guide/Manual_calibrations_SGAT.csv")
head(calib.all_SGAT)
```

## 5. Checking outliers

Sometimes we can get extremely high or low values for some calibration parameters in some periods. We detect and remove them before calculating mean values per GLS.

```{r check_out}
vars <- c(
  "zenith_PD1",     "zenith_PD2",     "zenith_PR3",
  "zenith0_PD1",    "zenith0_PD2",    "zenith0_PR3",
  "alpha.mean_PD1", "alpha.mean_PD2", "alpha.mean_PR3",
  "alpha.sd_PD1",   "alpha.sd_PD2",   "alpha.sd_PR3"
)

prd <- c("Migrate Technology", "Biotrack")

calib_list <- list()

for (p in prd) {
  cal.df <- calib.all_SGAT %>% filter(Producer == p)
  cal.df <- calib_out(cal.df, vars)
  calib_list[[p]] <- cal.df
}

calib.all <- bind_rows(calib_list) %>% arrange(geoID)
```

## 6. Mean values of calibration parameters

Once outliers are removed we calculate mean values per GLS. When all three periods are available we use the pre-deployment value (most reliable); otherwise we average the available periods.

```{r mean_calib}
calib.all_SGAT2 <- calib.all %>%
  group_by(Year, Colony, Producer) %>%
  mutate(
    mean_zenith = ifelse(
      is.na(zenith_PD1) | is.na(zenith_PD2) | is.na(zenith_PR3),
      mean(c(zenith_PD1, zenith_PD2, zenith_PR3), na.rm = TRUE),
      zenith_PD1
    ),
    mean_zenith0 = ifelse(
      is.na(zenith0_PD1) | is.na(zenith0_PD2) | is.na(zenith0_PR3),
      mean(c(zenith0_PD1, zenith0_PD2, zenith0_PR3), na.rm = TRUE),
      zenith0_PD1
    ),
    mean_alpha.mean = ifelse(
      is.na(alpha.mean_PD1) | is.na(alpha.mean_PD2) | is.na(alpha.mean_PR3),
      mean(c(alpha.mean_PD1, alpha.mean_PD2, alpha.mean_PR3), na.rm = TRUE),
      alpha.mean_PD1
    ),
    mean_alpha.sd = ifelse(
      is.na(alpha.sd_PD1) | is.na(alpha.sd_PD2) | is.na(alpha.sd_PR3),
      mean(c(alpha.sd_PD1, alpha.sd_PD2, alpha.sd_PR3), na.rm = TRUE),
      alpha.sd_PD1
    )
  ) %>%
  filter(!duplicated(geoID))

# Diagnostic plots
par(mfrow = c(3, 2))
boxplot(calib.all_SGAT2[, 25:26], main = "Zenith angles")
boxplot(calib.all_SGAT2[, 27:28], main = "Alpha parameters")
hist(calib.all_SGAT2$mean_zenith,     main = "Mean zenith",     xlab = "")
hist(calib.all_SGAT2$mean_zenith0,    main = "Mean zenith0",    xlab = "")
hist(calib.all_SGAT2$mean_alpha.mean, main = "Mean alpha.mean", xlab = "")
hist(calib.all_SGAT2$mean_alpha.sd,   main = "Mean alpha.sd",   xlab = "")

write_csv(calib.all_SGAT2, file = 'output/Supervised_calibrations_SGAT.csv')
```

Finally, we summarise calibration parameters per colony, species and year. This is useful for GLS that were not calibrated or have an invalid calibration period (e.g., recorded during equinox, too short, or not annotated).

```{r calib_summary}
calib_SGAT_summary <- calib.all_SGAT2 %>%
  group_by(Colony, Year, Producer) %>%
  dplyr::summarise(
    n_geos          = n_distinct(geoID),
    mean_zenith     = mean(mean_zenith,     na.rm = TRUE),
    mean_zenith0    = mean(mean_zenith0,    na.rm = TRUE),
    mean_alpha.mean = mean(mean_alpha.mean, na.rm = TRUE),
    mean_alpha.sd   = mean(mean_alpha.sd,   na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(Year) %>%
  distinct()

write_csv(calib_SGAT_summary, file = 'output/Summary_SGAT_calibrations.csv')
```
