---
title: "GLS supervised calibration analyses"
author: "Diego Vicente-Sastre"
format: html
editor: visual
date-modified: today
#bibliography: references.bib
execute: 
  echo: true
  warning: false
  message: false
---

## 1. Introduction: why calibrate GLS?

Twilight events are transitions of light between day and night, and *vice versa*. To establish them, it is set a predetermined light intensity threshold that divides daytime from night, so sunrises and sunsets can be temporarily noted (threshold-based method). The threshold must consist in the minimum possible value that is consistently above the noise in the light intensity captured at night (Lisovski et al., 2019).

GLS may vary in the way they measure and capture light due to environmental reasons or accuracy errors. Thus, it is first necessary to carry out a calibration of these devices to relate the expected light levels to those observed and be able to extract the final positions with minimal error. We employ the rooftop calibration method (Lisovski and Hahn, 2012). Each GLS has at least two calibrations, one pre-deployment and one post-recovery, in stationary locations with a minimum of light pollution. These calibrations should last at least seven days each, with the devices placed over a flat elevated surface, considering there cannot be any element that shadows (e.g., from buildings or rock walls), and in two different geographical locations, if possible.

The sunrise and sunset times captured by the GLS are associated with those of the area where the calibration was carried out. This information must be manually reviewed to correct or remove misplaced twilight events due to false light peaks in non-transition hours. At the same time, points on the map move with the twilight correction until they are all adjusted to the correct location. These calibrations involve subjective assessment, but Bråthen et al. (2021) showed that it could be performed consistently and with a high repeatability.

## 2. Load packages and functions

Some packages require to be installed remotely from other repositories outside of CRAN. Following lines need to be run only the first time.

```{r install_pkgs, include = TRUE, eval= FALSE}

# Both packages allow us to download packages from external repositories as GitHub
install.packages("pacman")
install.packages("devtools")
library(devtools)
install.packages("remotes")
library(remotes)

#Installing packages from GitHub
install_github("SWotherspoon/SGAT")
install_github("SWotherspoon/BAStag")
install_github("bmcclintock/momentuHMM")
install_github("SLisovski/TwGeos")
install_github("SLisovski/GeoLight")

```

We load other required packages

```{r loading_func, include=TRUE}
pacman::p_load('devtools','remotes', # installing external packages
               'readxl', # opening Excels and csv
               'tidyverse',  #managing data
               'lubridate', # dealing with dates
               'evaluate', # check errors
               'maps', 'rworldmap', 'maptools', # creating map for Geolight
               'GeoLight', 'BAStag', 'TwGeos' # geolocators
               )

if ("purrr" %in% .packages()){
  detach("package:purrr", unload = TRUE)
}
# to avoid problems with map function

```

Lastly, we load own functions, specifically designed for working with GLS data.

```{r extra_func}
source("functions/01_FUNCTIONS_Supervised_GLS_calibrations.R") # simplify working with GLS data

```

## 3. Read metadata and define thresholds

```{r load_data, include=FALSE, echo=F, message=F, warning=F}
mdata <- read_csv('input/GLS.files.csv', col_names = TRUE)
head(mdata)
info <- read_csv('input/GLS.info.csv', col_names = TRUE)
head(info)
calib.info <- read_csv('input/Info.calibrations.csv')
head(calib.info)
```

**Metadata file** contains the files available for each GLS (i.e., light, wet-dry and SST data) and the format of each file, which depends on the producer.

**Calibrations information file** contains the specific information for each calibration period of each GLS. Each GLS can be calibrated more than once and in different places. Here is added the coordinates of each calibration locality.

Following values are limiting the area for map the raw points inferred from light data along calibrations period. The map will be a helping tool, but not the main one.

```{r set_map, include=FALSE, echo=F, message=F, warning=F}

threshold = 2 # defines since which value is night or daylight

xlim <- c(-20, 0) #longitude limits of the map
ylim <- c(15, 40) #latitude limits of the map
map <- rworldmap::getMap(resolution = "coarse") # will help you to detect erroneous twilitght events

```

**Threshold** defines the light levels above which it is day and below it is night .Normally it is between 1.5 and 2.

The **map** will be a helping tool, but not the main reference to take decisions.

## 4. Loop to determine sun angles parameters

We are usually going to work with multiple GLSs at the same time. Due to that, the best option is apply a **loop** to chain each calibration with the following one. In that way, we will save time, avoiding manually introduce each values.

## 4. Extraction of calibration parameters

### 4.1. Read GLS information

We get all the **essential information** of the GLS.

```{r GLS_info}

i=1
gls.dir <-paste0(getwd(),'/input/GLS/')

# Load .lux file and extract the GLS id from the name of the file

info<-mdata[i,]
lightfile <- if_else(!is.na(info$File.adjlux), info$File.adjlux, info$File.lux)
lightfile_c <- paste0(gls.dir, lightfile)

# if (!file.exists(lightfile) | is.na(lightfile_c)) next

geoID <- info$geo_alfanum # Which GLS is processing

LoggerModel <- info$Modelo # logger model
Producer <- info$Producer # e.g "Migrate Technology","BAS","Biotrack","Lotek"



# Get info of the GLS and its calibration
calib<-calib.info%>%filter(geo_alfanum == geoID)# calibration info

```

Additionally, we can extract **other extra information** useful, for example for saving the files or if we are analyzing GLS from different colonies and species.

```{r GLS_extra_info}

# Additional information

col.name <- info$Colony
start_date<-as.Date(info$start_date, tz = 'GMT') #date start the trip
end_date <-as.Date(info$end_date, tz = 'GMT') #date end the trip
sp <- info$Sp # species
year_rec <- info$Year_rec # recovery year

cat('Processing geo (', i,') ', geoID, 'for year', year_rec, '\n')

```

### 4.2. Read light data

With the information read, we can use the function to read raw light data from any model of GLS. We need to specify the company o produce (e.g., Migrate Technology, Biotrack...), the ID of the GLS, the file/path that contains the data. Furthermore, we can subset the starting and ending date of recording data (i.e., deployment and recovery dates) if we are interested in.

In very few cases, some files from Migrate Technology cannot be read by *readMTlux* function. We can include them in *MTexcep* to read them correctly.

```{r read_light}
lu <- read_lightGLS(producer = Producer, ID = geoID, 
                    file = lightfile_c 
                    # start_date 
                    # end_date
                    )

```

### 4.3. Calibration parameters

For each calibration period we are going to extract several parameters that we will need for further steps processing GLS data. We base the loop procedure in `thresholdCalibration` function from `TwGeos` package.

-   **Zenith angles**:

    -   Zero deviation zenith angle

    -   Median deviation zenith angle

-   **Parameters of error distribution** (alpha parameters**:**

    -   Shape

    -   Scale

```{r main_loop, eval= FALSE}
twl_error.list <-list()
for (step in 1:3) {
  # step = 1
  
  # Define the calib_start and calib_end for the current step
  
  calib_start <- as.POSIXct(calib[[paste0("calib_start", step)]], tz = "GMT")
  calib_end <- as.POSIXct(calib[[paste0("calib_end", step)]]+1, tz = "GMT")# add one day to take all the hole calibration period
  n_days <- round(as.numeric(difftime(calib_end, calib_start, 
                                      units = 'days')))
  loc <- calib[[paste0("loc_calib", step)]]
  lon <- calib[[paste0("Lon", step)]]
  lat <- calib[[paste0("Lat", step)]]
  
  calib_SGAT <- NA
  
  if (!is.na(loc) && !is.na(calib_end)) {

    # Perform calibration for the current step
    twl <- calib_period(data = lu,
                        start = calib_start, 
                        end = calib_end, 
                        lon = lon, lat = lat,
                        threshold = threshold)
       # Optionally: Save light file of the calibration period
    trn <- twl %>%
      mutate(date_time = Twilight,
             type = ifelse(Rise == 'TRUE', 'Sunrise', 'Sunset'),
             confidence = Marker) %>%
      dplyr::select(date_time, type, confidence)
    
    if(step %in% c(1,2)) step <- 'predep'
    if(step %in% c(3,4)) step <- 'postrec'
  
    file <- paste(geoID, sp, year_rec, colony, step, '.trn', sep = '_')
    if(!dir.exists('output/calibrations')){
      dir.create('output/calibrations', recursive = TRUE, showWarnings = FALSE)
    }
      
    write.table(trn, file = paste0('output/calibrations/', file), 
                sep = ',', col.names = FALSE, row.names = FALSE, quote = FALSE)
    cat('Trn saved for calibration period', step, '\n')
    
    # Perform SGAT calibration
    calib_SGAT <- try(thresholdCalibration(twl$Twilight, twl$Rise, lon, lat, method = method))
    if (is.error(calib_SGAT)) {
      print('Error SGAT calibration')
      calib_SGAT <- c('NA', 'NA', 'NA', 'NA')
    } 
  } else {
    message('No calibration information for step',step,'\n')
    calib_SGAT <- c('NA', 'NA', 'NA', 'NA')
  }
  
  if(step == 3){cat('Geo done')}
  
  twl_error <- bind_cols(geoID = geoID, Producer = Producer, Year = year_rec, Colony = col.name,
                         n_days = n_days,
                         zenith = calib_SGAT[1], 
                         zenith0 = calib_SGAT[2], 
                         alpha.mean = calib_SGAT[3], 
                         alpha.sd = calib_SGAT[4],
                         Loc = loc)
  
  colnames(twl_error)[5:10]<- paste0(colnames(twl_error[5:10]),'_PD', step)
  
  
  twl_error.list[[step]] <-list(twl_error)
}


twl_error_GLS<- bind_rows(calib_SGAT.list)%>%
    group_by(geoID, Producer, Year, Colony) %>% 
    summarise_all(funs(na.omit(.)[1]))
  
  calib.list_SGAT[i]<-list(twl_error_GLS)
  

# Outside the loop joining SGAT calibration parameters
  
calib.all_SGAT <- bind_rows(calib.list_SGAT)

str(calib.all_SGAT)# check the structure of our data
calib.all_SGAT<-calib.all_SGAT%>%
  mutate_at(c(3,5:8,10:13,15:18), as.numeric)
 
```

```{r calib_results, include = FALSE}

# df <- read_csv("users_guide/Manual_calibrations_SGAT.csv")
# calib.all_SGAT <- df %>%
#   filter(geo_alfanum %in% c('11001', '12001', '19001', '33001', '41001', '42002', '47001'))%>%
#   dplyr::rename(Year = year)
# write_csv(calib.all_SGAT, file = "users_guide/Manual_calibrations_SGAT.csv")

calib.all_SGAT <- read_csv("users_guide/Manual_calibrations_SGAT.csv")
head(calib.all_SGAT)
```

## 5. Checking outliers

Sometimes we can get values extremely high or low of some of the calibration parameters in some of the calibration periods. To avoid issues, we will find the outliers and remove them to calculate mean values per GLS.

```{r check_out, echo=TRUE, include=TRUE}
# Define variables and producers to check
vars <- c("zenith_PD1", "zenith_PD2", "zenith_PR3", 
          "zenith0_PD1", "zenith0_PD2", "zenith0_PR3", 
          "alpha.mean_PD1", "alpha.mean_PD2", "alpha.mean_PR3", 
          "alpha.sd_PD1", "alpha.sd_PD2", "alpha.sd_PR3")

prd <- c("Migrate Technology", "Biotrack")

calib_list <- list()

# Loop through producers and apply the outlier detection and replacement
for (p in prd) {
  cal.df <- calib.all_SGAT %>%
    filter(Producer == p)
  cal.df <- calib_out(cal.df, vars)
  calib_list[[p]] <- cal.df
}

# Combine data from different producers
calib.all <- bind_rows(calib_list)

# Sort the data by geo_alfanum
calib.all <- calib.all %>% arrange(geo_alfanum)
# head(calib.all)

```

## 6. Mean values of calibration parameters

Once we have removed the outliers we can calculate mean values.

```{r mean_calib}
calib.all_SGAT2<- calib.all %>%
  group_by(Year, Colony, Producer)%>%
  mutate(mean_zenith = ifelse(is.na(zenith_PD1) | is.na(zenith_PD2) | is.na(zenith_PR3),
                              mean(c(zenith_PD1, zenith_PD2, zenith_PR3), na.rm = TRUE),
                              zenith_PD1))%>%
  mutate(mean_zenith0 = ifelse(is.na(zenith0_PD1) | is.na(zenith0_PD2) | is.na(zenith0_PR3),
                               mean(c(zenith0_PD1, zenith0_PD2, zenith0_PR3), na.rm = TRUE),
                               zenith0_PD1))%>%
  mutate(mean_alpha.mean = ifelse(is.na(alpha.mean_PD1) | is.na(alpha.mean_PD2) | is.na(alpha.mean_PR3),
                                  mean(c(alpha.mean_PD1, alpha.mean_PD2, alpha.mean_PR3), na.rm = TRUE),
                                  alpha.mean_PD1)) %>%
  mutate(mean_alpha.sd = ifelse(is.na(alpha.sd_PD1) | is.na(alpha.sd_PD2) | is.na(alpha.sd_PR3),
                                mean(c(alpha.sd_PD1, alpha.sd_PD2, alpha.sd_PR3), na.rm = TRUE),
                                alpha.sd_PD1))%>%
  filter(!duplicated(geo_alfanum))



par(mfrow= c(3,2))
boxplot(calib.all_SGAT2[,25:26])
boxplot(calib.all_SGAT2[,27:28])
hist(calib.all_SGAT2$mean_zenith)
hist(calib.all_SGAT2$mean_zenith0)
hist(calib.all_SGAT2$mean_alpha.mean)
hist(calib.all_SGAT2$mean_alpha.sd)

write_csv(calib.all_SGAT2, file = 'output/Supervised_calibrations_SGAT.csv')
```

Finally, we can summarize the calibration parameters per colony, species and year. This step will be helpful to analyse GLS than has not been calibrated or the calibration period is invalid (e.g., along equinox periods, too short period or not annotated).

```{r}
calib_SGAT_summary <- calib.all_SGAT2 %>%
  group_by(Colony, Year, Producer) %>%
  dplyr::summarise(
    n_geos = n_distinct(geo_alfanum),
    mean_zenith = mean(mean_zenith, na.rm = TRUE),
    mean_zenith0 = mean(mean_zenith0, na.rm = TRUE),
    mean_alpha.mean = mean(mean_alpha.mean, na.rm = TRUE),
    mean_alpha.sd = mean(mean_alpha.sd, na.rm = TRUE)
  ) %>%
  arrange(Year) %>%
  distinct()

write_csv(calib_SGAT_summary, file = 'output/Summary_SGAT_calibrations.csv')
```
