---
title: "GLS supervised calibrations analyses"
author: Diego Vicente Sastre
date: "`r Sys.Date()`"
output: github_document
abstract:
  "R script to extract sun angles of calibration periods of light-level
   geolocators"
---

**1. Load packages and functions**

Some packages require to be installed remotely from other repositories outside of CRAN. Following lines need to be run only the first time.

```{r setup, include=FALSE}
#install.packages("devtools")
#library(devtools)
#install.packages("remotes")
#library(remotes)
#install_github("SWotherspoon/SGAT")
#install_github("SWotherspoon/BAStag")
#install_github("bmcclintock/momentuHMM")
#install_github("SLisovski/TwGeos")
#install_github("SLisovski/GeoLight")
```

We load required packages and functions elaborated for simplify the process

```{r setup, include=FALSE}
pacman::p_load('devtools','remotes', # loading packages
               'readxl', # opening Excels and csv
               'tidyverse',  #manging data
               'lubridate', # dealing with dates
               'maps', 'rworldmap', 'maptools', # creating map 
               'GeoLight', 'BAStag', 'TwGeos','SGAT' # geolocators
               )


detach("package:purrr", unload = TRUE)
source('Functions/FUNCTION_datetime_conversion.R')
source("Functions/01_FUNCTIONS_Supervised_GLS_calibrations.R")
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff = 40), tidy = TRUE)
```

**2. Read information of each GLS and metadata**

```{r, include=FALSE, echo=F, message=F, warning=F}
mdata <- read_csv('RawData/GLS.files.csv', col_names = TRUE)
info <- read_csv('RawData/GLS.info.csv', col_names = TRUE)
calib.info <- read_csv('RawData/Info.calibrations.csv')

col.name <- 'Veneguera'
col.lat <- 27.84444
col.lon <- -15.78861

threshold = 2 # defones since which value is night or daylight

knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff = 40), tidy = TRUE)
```

Threshold defines the light levels above which it is day and below it is night .Normally it is between 1.5 and 2.

Following values are limiting the area for map the raw points inferred from light data along calibrations period. The map will be a helping tool, but not the main one.

```{r, include=FALSE, echo=F, message=F, warning=F}
xlim <- c(-20, 0)
ylim <- c(15, 40)
map <- rworldmap::getMap(resolution = "coarse")

knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff = 40), tidy = TRUE)
```

**3. Loop to determine sun angles parameters**

```{r, include=FALSE, echo=F, message=F, warning=F}

for (i in 1:nrow(GLS.files)) {
  #i=1
  setwd(paste0(wd, '/RawData/GLS/lux/'))
  
  # Load .lux file and extract the GLS id from the name of the file
  info<-GLS.files[i,]
  geoID <- info$geo_alfanum#Which GLS is processing
  
  lightfile <- if_else(!is.na(info$File.adjlux), info$File.adjlux, info$File.lux)
  if (!file.exists(lightfile) || geoID == GLS.files[i - 1, 'geo_alfanum']) {
    message(' Skipping geo (',i,') ', geoID, ' as there is no file or it has already been calibrated.\n')
    next
  }
  
  geoID <- info$geo_alfanum#Which GLS is processing
  year <- info$Year_rec
  cat('Processing geo (', i,') ', geoID, 'for year', year, '\n')
  
  # Select colony and determine coordinates
  sp <- info$Sp
  
  
  # Get info of the GLS and its calibration

  calib<-calib.info%>%filter(geo_alfanum == geoID)# calibration info
  
  start_date<-as.Date(info$start_date, tz = 'GMT') #date start the trip
  end_date <-as.Date(info$end_date, tz = 'GMT') #date end the trip
  
  LoggerModel<-info$Modelo#logger model
  Producer<- info$Producer #e.g "Migrate Technology","BAS","Biotrack","Lotek"

  
  lu <- read_lightGLS(producer = Producer, geo_trip, 
                      file_name = lightfile, 
                      start_date = start_date, 
                      end_date = end_date, 
                      MTexcep = MTlux_exception)
  
  # d.lux$Date <- as.POSIXct(strptime(d.lux$Date, format = '%Y-%m-%d %H:%M:%S', tz = 'GMT'))

  # Calculate zenith angle and alpha parameter from calibration periods. 
  # Our devices have been calibrated as maxim three times, so we repeat the loop 
  # three times
  
  twl_error.list <-list()
  for (step in 1:3) {
  
    # Define the calib_start and calib_end for the current step
    calib_start <- calib[[paste0("calib_start", step)]]
    calib_end <- calib[[paste0("calib_end", step)]] + 1 * 60 * 60 * 24
    n_days <- round(as.numeric(difftime(calib_end, calib_start, 
                                        units = 'days')))
    loc <- calib[[paste0("loc_calib", step)]]
    lon <- calib[[paste0("Lon", step)]]
    lat <- calib[[paste0("Lat", step)]]
    
    calib_SGAT <- NA
    
    if (!is.na(loc) && !is.na(calib_end)) {
      # Perform calibration for the current step
      twl <- calib_period(calib_start = calib_start, 
                          calib_end = calib_end, 
                          lon = lon, lat = lat,
                          threshold =threshold)
      
      # Save calibration data
      save_calib(twl, geoID, sp, year, col.name, step)
      
      # Perform SGAT calibration
      calib_SGAT <- calibration_SGAT(twl, lon = lon, lat = lat, method = 'gamma')
      
    } else {
      message('No calibration information for step',step,'\n')
      calib_SGAT <- c('NA', 'NA', 'NA', 'NA')
    }
    
    if(step == 3){cat('Geo done')}
    
    twl_error <- bind_cols(geoID = geoID, Producer = Producer, Year = year, Colony = col.name,
                           n_days = n_days,
                           zenith = calib_SGAT[1], 
                           zenith0 = calib_SGAT[2], 
                           alpha.mean = calib_SGAT[3], 
                           alpha.sd = calib_SGAT[4],
                           Loc = loc)
   
    colnames(twl_error)[5:10]<- paste0(colnames(twl_error[5:10]),'_PD', step)
    
    
    twl_error.list[[step]] <-list(twl_error)
  }
  twl_error_GLS<- bind_rows(calib_SGAT.list)%>%
    group_by(geoID, Producer, year, Colony) %>% 
    summarise_all(funs(na.omit(.)[1]))
  
  calib.list_SGAT[i]<-list(twl_error_GLS)
  
}
```
