---
title: "Trn automated calculation"
author: "Diego Vicente-Sastre"
format: html
editor: visual
date-modified: today
# bibliography: references.bib
execute: 
  echo: true
  warning: false
  message: false
---

## 1. Introduction: how to get positions from light data?

Light-level geolocators (GLS) record ambient light intensity at fixed intervals throughout the deployment period. From these raw light measurements, it is possible to infer the timing of sunrise and sunset — so-called twilight events — which are then used to estimate geographic positions based on the time of local noon (longitude) and day length (latitude).

This script processes the raw light data for each GLS into a set of twilight events saved as `.trn` files. These files are the main input for the SGAT location modelling in the next step. The automated procedure uses three main steps: (1) detection of twilight events with `twilightCalc` from the `GeoLight` package, (2) removal of unrealistic events with `twilight_cleanup`, and (3) adjustment of slightly displaced events with `move_twilights`.

## 2. Load packages and functions

All packages needed here were installed in script 01. We only need to load them.

```{r loading_func}
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")

pacman::p_load(
  'readxl',                       # opening Excels and csv
  'tidyverse',                    # managing data
  'lubridate',                    # dealing with dates
  'evaluate',                     # check errors
  'GeoLight', 'BAStag', 'TwGeos', # geolocators
  'suncalc'                       # twilight times by location
)

# Detach purrr to avoid conflicts with base map()
if ("purrr" %in% .packages()) {
  detach("package:purrr", unload = TRUE)
}
```

We also load custom functions designed for working with GLS data.

```{r extra_func}
source('functions/FUNCTION_twilight_cleanup.R')
source('functions/FUNCTION_move_twilights.R')
source('functions/01_FUNCTIONS_GLS_analysis.R')
```

## 3. Read metadata and calibration parameters

```{r load_data}
mdata     <- read_csv('input/GLS.files.csv',                     col_names = TRUE)
calibs    <- read_csv('output/Supervised_calibrations_SGAT.csv',  col_names = TRUE)
calib.sum <- read_csv('output/Summary_SGAT_calibrations.csv',     col_names = TRUE)
```

**Metadata file** contains the files available for each GLS (light, wet-dry and SST data) and the format of each file depending on the producer.

**Calibrations data** contains the zenith angles and the shape and scale parameters of the twilight error distribution, derived from `thresholdCalibration` in script 01.

**Summary of calibrations** contains mean values per year, colony and producer, used for GLS without individual calibration.

## 4. Processing raw light data to twilight events

As in the previous script, we work through a **loop** over all GLS. The single-GLS example below (with `i = 1`) shows each step in detail; in a full analysis, wrap everything from section 4.1 to 4.5 in `for (i in seq_len(nrow(mdata)))`.

### 4.1. Read GLS information

```{r GLS_info}
gls.dir <- paste0(getwd(), '/input/GLS/')
i <- 1

info        <- mdata[i, ]
lightfile   <- if_else(!is.na(info$File.adjlux), info$File.adjlux, info$File.lux)
lightfile_c <- paste0(gls.dir, lightfile)

geoID    <- info$geo_alfanum
geo_trip <- info$geo_trip
Producer <- info$Producer
col.name <- info$Colony
col_lat  <- info$Colony.lat
col_lon  <- info$Colony.lon

start_date <- as.Date(info$start_date, tz = 'GMT')
end_date   <- as.Date(info$end_date,   tz = 'GMT')
year_rec   <- info$Year_rec

cat('Processing geo (', i, ')', geo_trip, 'for year', year_rec, '\n')
```

The function `select_sunAngle` selects the zenith angle from the individual calibration. If the GLS was not calibrated or is not in the calibration dataset, parameters are taken from the colony/year summary.

```{r sun_angle}
sun_angle <- select_sunAngle(
  producer  = Producer,
  geoID     = geoID,
  colony    = col.name,
  year      = year_rec,
  calib.sum = calib.sum,
  calibs    = calibs,
  param     = FALSE
)
```

### 4.2. Read light data

We read the raw light data subset to the deployment and recovery dates. We then derive the recording interval, the start and end years, and the equinox dates for the trip.

```{r read_light}
lu <- read_lightGLS(
  producer   = Producer,
  ID         = geoID,
  file       = lightfile_c,
  start = start_date,
  end   = end_date
)

if (is.null(lu) || nrow(lu) == 0) stop('No light data for geo ', geoID)

# Recording interval in minutes
li <- median(difftime(lu$dtime[2:nrow(lu)], lu$dtime[1:(nrow(lu) - 1)], units = 'mins'))

# Start and end year of the recording
start_year <- year(lu[1, 1])
end_year   <- year(lu[nrow(lu), 1])

if (start_year == end_year)
  stop('Geo ', geo_trip, ' has less than one year of data — remove from dataset')

# Equinox dates for this trip
Sep_eq <- as.Date(paste0(start_year, '-09-21'))
Mar_eq <- as.Date(paste0(end_year,   '-03-21'))
```

### 4.3. Automated calculation of twilight events

Twilight events are detected with `twilightCalc` from `GeoLight`. Two custom SEATRACK functions then refine the results: `twilight_cleanup` removes unrealistic events, and `move_twilights` corrects events that are slightly displaced relative to the running mean.

```{r twilight_calc}
twl <- try(
  twilightCalc(lu$dtime, lu$lux,
               ask            = FALSE,
               preSelection   = TRUE,
               LightThreshold = 2,
               maxLight       = li),
  silent = TRUE
)

if (inherits(twl, 'try-error'))
  stop('Twilight calculation failed for geo ', geo_trip)

# Fix type=1/2 misassignments (two consecutive same-type events)
for (j in 1:(nrow(twl) - 2)) {
  if (sum(twl$type[j:(j + 2)]) == 3) twl$type[j + 1] <- 2
  if (sum(twl$type[j:(j + 2)]) == 6) twl$type[j + 1] <- 1
}

# GeoLight removes one extra minute vs BAStag/intiproc — compensate
twl$tFirst[twl$type  == 2] <- twl$tFirst[twl$type  == 2] + 60
twl$tSecond[twl$type == 1] <- twl$tSecond[twl$type == 1] + 60

twl <- twl[, c(1:3)]
```

```{r twilight_clean}
par(mfrow = c(2, 1))

twl2 <- twilight_cleanup(
  df                          = twl,
  breedingloc_lon             = col_lon,
  breedingloc_lat             = col_lat,
  months_breeding             = c(5, 6, 7, 8),
  months_extensive_legtucking = NA,
  show_plot                   = TRUE
)
twl2 <- distinct(twl2)

twl3 <- move_twilights(
  df                = twl2,
  minutes_different = 15,
  sun               = sun_angle,
  show_plot         = TRUE
)
twl3 <- distinct(twl3)
```

Optionally, additional filters can be applied. They are calculated but not applied by default — uncomment the `filter` lines to use them.

```{r optional_filters}
twl3 <- twl3 %>%
  mutate(
    loess        = loessFilter(twl3, k = 3, plot = TRUE),
    speed_filter = distanceFilter(twl3, degElevation = sun_angle,
                                  distance = 500, units = "hour")
  )
# Uncomment to apply:
# twl3 <- twl3 %>% filter(loess == TRUE)
# twl3 <- twl3 %>% filter(speed_filter == TRUE)
```

### 4.4. Assign confidence values

Each twilight event receives a confidence value of 9. Events close to the equinox periods receive lower values (2 or 1) because positions are unreliable near equinoxes.

```{r confidence}
twl4 <- twl3 %>%
  mutate(
    Twilight   = tFirst,
    Rise       = ifelse(type == '1', TRUE, FALSE),
    confidence = 9,
    Julian     = yday(Twilight)
  )

twl4 <- twl4 %>%
  mutate(
    confidence = case_when(
      between(Twilight, Sep_eq - 20, Sep_eq + 20) ~ 2,
      between(Twilight, Sep_eq - 10, Sep_eq + 10) ~ 1,
      between(Twilight, Mar_eq - 20, Mar_eq + 20) ~ 2,
      between(Twilight, Mar_eq - 10, Mar_eq + 10) ~ 1,
      TRUE ~ confidence
    )
  )
```

### 4.5. Save .trn files

We save two formats: the main `.trn` file used by subsequent scripts, and a backup in the format readable by the Locator (BirdTracker) software.

```{r save_trn}
trn <- twl4[, c(1:3, 6, 7, 9, 8, 4, 5)]

file.trn <- paste0(geo_trip, '_', info$Sp, '_', year_rec, '_', info$Colony, '_automated.trn')

if (!dir.exists('output/trn'))
  dir.create('output/trn', recursive = TRUE, showWarnings = FALSE)

write.table(trn,
            file      = paste0('output/trn/', file.trn),
            sep       = ',',
            col.names = TRUE,
            row.names = FALSE,
            quote     = FALSE)

# Backup format for Locator (BirdTracker) software
trn.bak <- trn %>%
  mutate(
    date_time = format(Twilight, format = '%d/%m/%y %H:%M:%S'),
    type      = ifelse(Rise == TRUE, 'Sunrise', 'Sunset')
  ) %>%
  arrange(date_time) %>%
  dplyr::select(date_time, type, confidence)

file.trn.bak <- paste0(geo_trip, '_', info$Sp, '_', year_rec, '_', info$Colony, '.trn')

if (!dir.exists('output/trn_bak'))
  dir.create('output/trn_bak', recursive = TRUE, showWarnings = FALSE)

write.table(trn.bak,
            file      = paste0('output/trn_bak/', file.trn.bak),
            sep       = ',',
            col.names = FALSE,
            row.names = FALSE,
            quote     = FALSE)

cat('Trn from geo (', i, ')', geo_trip, 'saved\n')

# Update metadata with output file name
mdata[i, c(21:23)] <- list(paste0('trn/', file.trn), 'Data', 'No')
write_csv(mdata, file = 'input/GLS.files2.csv')
```
