---
title: "Trn automated calculation"
author: "Diego Vicente-Sastre"
format: html
editor: visual
date-modified: today
# bibliography: references.bib
execute: 
  echo: true
  warning: false
  message: false
---

## 1. Introduction: how to get positions from light data?

## 2. Load packages and functions

All packages we need for the analyses have been installed in the previous script. At this step, we only need to load them.

```{r loading_func, include=TRUE}
# Installing all the packages and load their libraries we need to run the script

pacman::p_load('readxl', # opening Excels and csv
               'tidyverse',  #manging data
               'lubridate', # dealing with dates
               'evaluate', # check errors
               'GeoLight', 'BAStag', 'TwGeos',# geolocators
               'suncalc' # knowing twilights depending on the place
               )

```

We also load own functions, specifically designed for working with GLS data.

```{r extra_func}

source('functions/FUNCTION_twilight_cleanup.R')
source('functions/FUNCTION_move_twilights.R')

source('functions/01_FUNCTIONS_GLS_analysis.R')
```

## 3. Read metadata and calibration parameters

```{r load_data, include=FALSE, echo=F, message=F, warning=F}

mdata <- read_csv('input/GLS.files.csv', col_names = TRUE)
head(mdata)
calibs <- read_csv('output/Supervised_calibrations_SGAT.csv')
head(calibs)
calib.sum <- read.csv('output/Summary_SGAT_calibrations.csv')
head(calib.sum)
```

**Metadata file** contains the files available for each GLS (i.e., light, wet-dry and SST data) and the format of each file, which depends on the producer.

**Calibrations data** contains the zenith angles and shape and scale of the distribution of the twilight estimation error. Both derived from the `thresholdCalibration` function from `TwGeos` package

In addition, mean values per year, colony and producer are stored in **summary of calibrations**

## 4. Processing raw light data to twilight events data

As in the previous step, we will work through a **loop** to process all the GLS files.

### 4.1. Read GLS information

We read the metadata information of each GLS as in the previous script.

```{r}

gls.dir <-paste0(getwd(),'/input/GLS/')

i=1

# Load .lux file and extract the GLS id from the name of the file
info<-mdata[i,]

N_trip <- info$N_trips
lightfile <- if_else(!is.na(info$File.adjlux), info$File.adjlux, info$File.lux)
lightfile_c <- paste0(gls.dir, lightfile)

# if (!file.exists(lightfile) | is.na(lightfile_c)){
#   mdata[i, c(21:23)] <- list(NA, 'No data', 'No')
#   next
# }

geoID<-info$geo_alfanum # Which GLS is processing
geo_trip<-info$geo_trip # Which GLS trip is selecting

LoggerModel <- info$Modelo # logger model
Producer <- info$Producer # e.g "Migrate Technology","BAS","Biotrack","Lotek"

start_date<-as.Date(info$start_date, tz = 'GMT') #date start the trip
end_date <-as.Date(info$end_date, tz = 'GMT') #date end the trip
year_rec <- info$Year_rec #which year we recovered the GLS

col.name <- info$Colony
col_lat <- info$Colony.lat
col_lon <- info$Colony.lon

sun_angle <- select_sunAngle(producer = Producer, col.name, year = year_rec,
                              calib.sum, calibs, param = F)
```

The function `select_sunAngle` allows user to choose the zenith angle determined during the calibration step If the GLS was not calibrated or was not in the dataset, the zenith angle and the twilight error parameters (alpha mean and alpha sd; also known as shape and scale) are selected from the summary dataset. The function will select the parameters according to the producer, colony and year of the GLS trip.

### 4.2. Read light data

We read raw light data as in the previous step. As difference, we determine deploying and recovering dates of the GLS. Then we define the starting and ending dates of data recorded to check the yearly data are completed. For next step, we define equinox dates of these years.

```{r}


lu <- read_lightGLS(producer = Producer, ID = geoID, 
                    file = lightfile_c,
                    start_date,
                    end_date,
                    )

# if (is.null(lu) | nrow(lu) == 0) {
#   mdata[i, c(21:23)] <- list(NA, 'No data', NA)
#   next
# }

#Determine the recording interval
li<-median(difftime(lu$dtime[2:nrow(lu)],lu$dtime[1:(nrow(lu)-1)],units='mins'))

# Calculate the start and end year
start_year <- year(lu[1, 1])
end_year <- year(lu[nrow(lu), 1])

# Check if the GLS trip contains enough data for analysis
if (start_year == end_year) {
  # Not enough data, indicate for removal
  GLS.files2[i, c(21:23)] <- list(NA, 'Data', 'Remove')
  message(paste0('Remove ', geo_trip, ' without enough data'))
  #next
}

# Select the equinoxes of this year
Sep_eq <- as.Date(paste0(start_year, '-09-21'))
Mar_eq <- as.Date(paste0(end_year, '-03-21'))
```

### 4.3. Automated calculation of twilight events

This is the main step of this part of the GLS analysis. Twilight events (sunrise and sunset) times are calculated using the `twilightCalc` function from `GeoLight` package. Then code is based on two functions designed by SEATRACK project: `twilight_cleanup` and `move_twilights` functions. The first one is used to remove the erroneous twilight events and the second one to move the twilight events that are slightly inaccurate (e.g., earlier or later than the running mean value of the next ones).

```{r}
# Logaritmic values can help and are recommended in some cases
# lu$lux2<-log(lu$lux)


# This function calculates automatically twl for all the period of light data
twl <- try(
  twilightCalc(lu$dtime, lu$lux, 
               ask = FALSE, preSelection = TRUE, 
               LightThreshold = 2, maxLight = li),
  silent = TRUE
)

# if (inherits(twl, 'try-error')) {
#   message(paste('Geo (', i, ') ', geo_trip,' wihtout twilight calculation - Skipped'))
#   mdata[i, c(21:23)] <- list(NA, "Data. Problem with twilight calculation", NA)
#   next
# }


# type=1 and type=2 not always correctly assigned, this fixes most of these cases:
for(j in 1:(nrow(twl)-2)){
  if(sum(twl$type[j:(j+2)])==3){twl$type[j+1]<-2}
  if(sum(twl$type[j:(j+2)])==6){twl$type[j+1]<-1}}


# Geolight is removing a minute extra unlike bastrack and intiproc when advancing sunset, 
# I add a minute to compensate

twl$tFirst[twl$type == 2] <- twl$tFirst[twl$type == 2] + 60  
twl$tSecond[twl$type == 1] <- twl$tSecond[twl$type ==   1] + 60

twl<-twl[,c(1:3)]

# Open a pdf to record the plots that illustrate the edition of twilights
# pdf(file= paste0(wd,'/Plots/twlEdit/', geo_trip,'_', year_rec, '_', colony,'.pdf'))
par(mfrow= c(2,1))

# 4.2. Remove and clean unrealistic twilight ####
twl2<-twilight_cleanup(df = twl,
                       breedingloc_lon = col_lon,
                       breedingloc_lat = col_lat,
                       months_breeding = c(5,6,7,8), 
                       months_extensive_legtucking = NA, 
                       show_plot = T)
twl2<-distinct(twl2)

twl3<-move_twilights(df = twl2,
                     minutes_different= 15,
                     sun = sun_angle,
                     show_plot = TRUE)
twl3<-distinct(twl3)

# Optional filters


twl3<-twl3%>%
  mutate(loess = loessFilter(twl3, k=3, plot = T),
         speed_filter = distanceFilter(twl3, degElevation = sun_angle, distance = 500, 
                                       units = "hour"))

#filter(loess == TRUE)%>%
#filter(speed_filter == TRUE)
```

### 4.4. Save trn files

In order to save the data as .trn files, we create a column to give a confidence value to each twilight event. Maximum confidence value is 9, which is given to every twilight with this procedure except to those twilight events close to the equinox dates.

We save the files with two formats, one as backup that has the same format of the .trn files that can be obtained using the Locator software. The other is more practical for our purposes.

```{r}
# Adding some columns to read it easily by any algorithm functions
twl4<-twl3%>%
  mutate(Twilight = tFirst,
         Rise = ifelse(type=='1',TRUE,FALSE),
         confidence = 9,
         Julian = yday(Twilight))

# Assign confidence for equinox periods as we used to do 
twl4 <- twl4 %>%
  mutate(
    confidence = case_when(
      between(Twilight, Sep_eq - 20, Sep_eq + 20) ~ 2,
      between(Twilight, Sep_eq - 10, Sep_eq + 10) ~ 1,
      between(Twilight, Mar_eq - 20, Mar_eq + 20) ~ 2,
      between(Twilight, Mar_eq - 10, Mar_eq + 10) ~ 1,
      TRUE ~ confidence
    )
  )

# Select columns to save
trn<- twl4[, c(1:3,6,7,9,8,4,5)]
# trn <- distinct(trn)

# Name of the file
file.trn<-paste0(geo_trip,'_', info$Sp,'_', year_rec,'_',info$Colony, '_automated.trn')

write.table(trn, 
            paste0('output/', file.trn), 
            sep = ',', col.names =  T, row.names = F, quote = F)


# Saving a trn file in old format to see the results in Locator (BirdTracker)
trn.bak <- trn %>%
  mutate(
    date_time = format(Twilight, format = '%d/%m/%y %H:%M:%S'),
    type = ifelse(Rise == T, 'Sunrise', 'Sunset')
  ) %>%
  arrange(date_time) %>%
  dplyr::select(date_time, type, confidence)

# Change the format of the date column to match with 'Locator' software

file.trn.bak<-paste0(geo_trip,'_', info$Sp,'_', year_rec,'_',info$Colony, '.trn')

if(!dir.exists('output/trn_bak')){
  dir.create('output/trn_bak', recursive = TRUE, showWarnings = FALSE)
}

#Save trn file
write.table(trn.bak, 
            paste0('output/trn_bak', file.trn.bak), 
            sep = ',', col.names =  F, row.names = F, quote = F)



# Add the name of the file to the dataset
cat('Trn from geo (', i,') ',geo_trip, ' is saved \n')
mdata[i, c(21:23)] <- list(file.trn, 'Data', 'No')


write_csv(mdata, file = 'input/GLS.files2.csv')
```
