---
title: "SGAT locations from trn files"
author: "Diego Vicente-Sastre"
format: html
editor: visual
date-modified: today
# bibliography: references.bib
execute: 
  echo: true
  warning: false
  message: false
---

## 1. Introduction: modelling GLS positions

Once twilight events have been calculated and cleaned (script 02), the next step is to convert them into geographic positions. We use the **threshold path method** implemented in the `SGAT` package (Wotherspoon et al.), which estimates the most likely longitude from the timing of local noon and the latitude from day length at each twilight pair.

Two key parameters control the model:

- **Zenith angle**: the solar angle at which the GLS records the transition between day and night, derived from the rooftop calibration (script 01).
- **tol**: a tolerance parameter that modulates the sensitivity of the latitude estimate around the equinox, where day length is uninformative. A separate tol value is calculated for the autumn and spring equinox periods.

Colony days information (from the wet-dry and light sensors) is then used to anchor positions at the known breeding location on days when the bird was confirmed to be at the colony.

## 2. Load packages and functions

```{r loading_func}
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")

pacman::p_load(
  'readxl',                            # opening Excels and csv
  'tidyverse',                         # managing data
  'lubridate',                         # dealing with dates
  'SGAT',                              # geolocator threshold path model
  'evaluate',                          # check errors
  'geosphere',                         # destPoint for random colony offsets
  'rnaturalearth', 'rnaturalearthdata', # world map for plots
  'ggplot2', 'cowplot', 'patchwork'    # plots
)

# Detach purrr to avoid conflicts with base map()
if ("purrr" %in% .packages()) {
  detach("package:purrr", unload = TRUE)
}
```

```{r extra_func}
source('functions/01_FUNCTIONS_GLS_analysis.R')
```

## 3. Read metadata and calibration parameters

```{r load_data}
mdata2    <- read_csv('input/GLS.files2.csv',                    col_names = TRUE)
calibs    <- read_csv('output/Supervised_calibrations_SGAT.csv', col_names = TRUE)
calib.sum <- read_csv('output/Summary_SGAT_calibrations.csv',    col_names = TRUE)
col.days  <- read_csv('input/colony_dates.csv')
```

Apart from the **metadata and calibration files**, we include a file with the dates **the animal was confirmed to be at the colony**. These dates are determined from two sources: nights when the GLS recorded no light (animal inside the burrow) and breeding-season days when the animal spent more than 75% of the night dry (close to the colony area).

## 4. Processing twilight events into tracking positions

As in previous scripts, the full analysis runs inside a loop over all GLS. The example below (with `i = 1`) shows each step in detail.

### 4.1. Read GLS information

```{r GLS_info}
i <- 1

info      <- mdata2[i, ]
trnfile   <- if_else(!is.na(info$File.trn), info$File.trn, NA_character_)
trnfile_c <- paste0('output/trn/', trnfile)

if (is.na(trnfile) || !file.exists(trnfile_c))
  stop('Trn file not found for geo ', info$geo_trip, ': ', trnfile_c)

ring     <- info$Ring
sp       <- info$Sp
geoID    <- info$geo_alfanum
geo_trip <- info$geo_trip
Producer <- info$Producer
col.name <- info$Colony
col_lat  <- info$Colony.lat
col_lon  <- info$Colony.lon
year_rec <- info$Year_rec

start_date <- as.Date(info$start_date, tz = 'GMT')
start_year <- year(start_date)
end_date   <- as.Date(info$end_date, tz = 'GMT')
end_year   <- year(end_date)

cat('Processing geo (', i, ')', geo_trip, 'for year', year_rec,
    'from', col.name, '\n')
```

We read the `.trn` file and subset to the deployment period. We also retrieve the calibration parameters for this GLS — individual calibration if available, colony/year summary otherwise.

```{r read_trn}
trn <- read_csv(trnfile_c, col_names = TRUE, show_col_types = FALSE)

twl <- trn %>%
  filter(Twilight >= start_date & Twilight <= end_date) %>%
  distinct(Twilight, .keep_all = TRUE) %>%
  arrange(Twilight)

if (nrow(twl) == 0) stop('No twilight events in date range for geo ', geo_trip)
```

```{r calib_params}
calib.param <- select_sunAngle(
  producer  = Producer,
  geoID     = geoID,
  colony    = col.name,
  year      = year_rec,
  calib.sum = calib.sum,
  calibs    = calibs,
  param     = TRUE
)

zenith  <- calib.param$zenith_mean
zenith0 <- calib.param$zenith0_mean
alpha   <- calib.param$alpha

cat('Zenith:', zenith, '| Zenith0:', zenith0, '\n')
```

### 4.2. Determine tol parameter

The `tol` parameter modulates the latitude tolerance around equinox periods, where GLS positions are unreliable. As Lisovski et al. (2019) describe, a larger tol allows more spread in latitude estimates, which is appropriate near the equinox. We calculate separate values for the autumn (`tol_sep`) and spring (`tol_mar`) equinoxes by evaluating the latitude spread across a range of tol values.

```{r calc_tol}
Sep_eq <- as.Date(paste0(start_year, '-09-21'))
Mar_eq <- as.Date(paste0(end_year,   '-03-21'))

tol_res <- calc_tol(
  twl    = twl,
  zenith = zenith,
  n      = 0.001,
  d      = 20,
  Sep_eq = Sep_eq,
  Mar_eq = Mar_eq,
  k1     = 0.95,
  k2     = 0.5,
  plot   = TRUE
)

tol_res$plot
```

### 4.3. SGAT initial path

If a valid spring tol was found, the trip is split at January 1st and each half is modelled with its corresponding tol value. Otherwise a single tol is applied to the full trip.

```{r initial_path}
if (!is.null(tol_res$tol_mar) && !is.na(tol_res$tol_mar)) {

  twl_1 <- twl %>% filter(Twilight >= start_date &
                             Twilight <= as.Date(paste0(end_year, '-01-01')))
  twl_2 <- twl %>% filter(Twilight >  as.Date(paste0(end_year, '-01-01')) &
                             Twilight <= end_date)

  path_1 <- thresholdPath(twl_1$Twilight, twl_1$Rise,
                           zenith = zenith, tol = tol_res$tol_sep)
  path_2 <- thresholdPath(twl_2$Twilight, twl_2$Rise,
                           zenith = zenith, tol = tol_res$tol_mar)

  x0   <- rbind(path_1$x, path_2$x)
  time <- c(path_1$time, path_2$time)

} else {
  path <- thresholdPath(twl$Twilight, twl$Rise,
                        zenith = zenith, tol = tol_res$tol_sep)
  x0   <- path$x
  time <- path$time
}

colnames(x0) <- c('lon', 'lat')
cat('Initial path determined\n')

p1 <- plot_SGAT(x0, time, d = 20, Sep_eq, Mar_eq, col_lat, col_lon)
p1
```

### 4.4. Save raw SGAT track

```{r save_raw}
raw.points <- cbind(time, as.data.frame(x0))
colnames(raw.points) <- c('Date_Time', 'lon', 'lat')

raw.points <- bind_cols(
  Ring       = ring,       Species    = sp,
  Colony     = col.name,   geoID      = geoID,
  geo_trip   = geo_trip,   Producer   = Producer,
  start_year = start_year, end_year   = end_year,
  tol_sep    = tol_res$tol_sep,
  tol_mar    = tol_res$tol_mar,
  raw.points
)

cat('Raw track saved for geo', geo_trip, '\n')
```

### 4.5. Clean SGAT positions through colony days information

Colony days are used to anchor positions to the known breeding location. Three levels of confidence are assigned:

- `fixedx = 1`: confirmed colony days (no light recorded) → position set exactly at colony
- `fixedx = 2`: probable colony days (dry night \> 75%) → position set at a random point within 100 km of the colony
- `fixedx = 3`: intermediate confidence → position left free

```{r colony_days}
cdays <- col.days %>%
  dplyr::filter(geo_trip == .env$geo_trip) %>%
  filter(light == TRUE | wetdry == TRUE)

auxi <- data.frame(date = as.Date(time)) %>%
  cbind(as.data.frame(x0), row = seq_len(nrow(x0))) %>%
  filter(date %in% cdays$date)

auxi <- left_join(auxi, cdays[, c(3, 6, 9)], by = 'date')

fixedx        <- rep(0, nrow(x0))
fixedx[auxi[is.na(auxi$wetdry),  ]$row] <- 3
fixedx[auxi[auxi$wetdry == TRUE, ]$row] <- 2
fixedx[auxi[auxi$light  == TRUE, ]$row] <- 1
fixedx[1:2] <- 1  # first two positions always at colony

# Last positions: colony if full deployment, free if resuscitated GLS
if (end_date == as.Date(twl$Twilight[nrow(twl)])) {
  fixedx[(nrow(x0) - 1):nrow(x0)] <- 1
} else {
  fixedx[(nrow(x0) - 1):nrow(x0)] <- 0
}

# Assign colony coordinates or a nearby random point
for (k in which(fixedx != 0)) {
  if (fixedx[k] == 1) {
    x0[k, 'lon'] <- col_lon
    x0[k, 'lat'] <- col_lat
  } else if (fixedx[k] == 2) {
    random_p     <- geosphere::destPoint(p = c(col_lon, col_lat),
                                         b = runif(1, 0, 360), d = 100 * 1000)
    x0[k, 'lon'] <- as.numeric(random_p[1])
    x0[k, 'lat'] <- as.numeric(random_p[2])
  }
}

p2 <- plot_SGAT(x0, time, d = 20, Sep_eq, Mar_eq, col_lat, col_lon)
p2
cat('Colony days corrected\n')
```

```{r save_corrected}
raw.points2 <- cbind(time, as.data.frame(x0))
colnames(raw.points2) <- c('Date_Time', 'lon', 'lat')

raw.points2 <- bind_cols(
  Ring       = ring,       Species    = sp,
  Colony     = col.name,   geoID      = geoID,
  geo_trip   = geo_trip,   Producer   = Producer,
  start_year = start_year, end_year   = end_year,
  tol_sep    = tol_res$tol_sep,
  tol_mar    = tol_res$tol_mar,
  raw.points2
)
```

## 5. Save SGAT raw tracks

After running the loop over all GLS, the results are saved. Two files are produced: one with the raw threshold path positions and one with colony-day corrections applied.

```{r save_all, eval=FALSE}
# ── Run after processing all GLS ────────────────────────────────────────────
raw_trips  <- bind_rows(raw_trips.list)
raw_trips2 <- bind_rows(raw_trips2.list)

if (!dir.exists('output')) dir.create('output', recursive = TRUE)

write_csv(raw_trips,  file = 'output/Raw_SGAT_tracks_tol.csv')
write_csv(raw_trips2, file = 'output/Raw_SGAT_tracks_tol_cdays.csv')
write_csv(mdata2,     file = 'input/GLS.files3.csv')

cat('Tracks saved to output/\n')
cat('Updated metadata saved to input/GLS.files3.csv\n')
```
