---
title: "SGAT locations from trn files"
author: "Diego Vicente-Sastre"
format: html
editor: visual
date-modified: today
# bibliography: references.bib
execute: 
  echo: true
  warning: false
  message: false
---

## 1. Introduction: modelling GLS positions

## 2. Load packages and functions

All packages we need for the analyses have been installed in the previous script. At this step, we only need to load them.

```{r loading_func, include=TRUE}
# Installing all the packages and load their libraries we need to run the script

pacman::p_load('readxl', # opening Excels and csv
               'tidyverse',  #manging data
               'lubridate', # dealing with dates
               'maps', 'rworldmap', 'maptools', # plot world maps
               'GeoLight', 'BAStag', 'TwGeos','SGAT', # geolocator data
               'evaluate', # check errors
               'ggplot2', 'viridis', 'cowplot', 'patchwork'# plots
)

```

We also load own functions, specifically designed for working with GLS data.

```{r}

source('functions/01_FUNCTIONS_Supervised_GLS_calibrations.R')

```

## 3. Read metadata and calibration parameters

```{r load_data, include=FALSE, echo=F, message=F, warning=F}

mdata2 <- read_csv('input/GLS.files2.csv', col_names = TRUE)
head(mdata2)
calibs <- read_csv('output/Supervised_calibrations_SGAT.csv')
head(calibs)
calib.sum <- read.csv('output/Summary_SGAT_calibrations.csv')
head(calib.sum)
col.days <- read_csv('input/colony_days.csv')
```

In this step we include apart the **metadata and calibrations files**, a file that contains the dates **the animal has been at the colony**. This dates has been determined regarding the days the GLS recorded no-light (i.e., the animal is inside the burrow) and the nights the animal spent more the 75% of the night in dry (i.e., animal was close enough to the colony area, flying and resting on land).

For more the details, see...

## 4. Processing twilight events data into tracking data

We will work again through a **loop** to process all the GLS.

### 4.1. Read GLS information

We load all the information we have load in previous scripts, adding the calibration parameters and reading the trn file with the twilight events data.

```{r}
i=1

info<-mdata2[i,]
trnfile <- if_else(!is.na(info$File.trn), info$File.trn, NA)
trnfile_c <- paste0('output/', trnfile)

# Trip information
ring<- info$Ring # individual
sp <- info$Sp # species
geoID<-info$geo_alfanum # Which GLS is processing
geo_trip<-info$geo_trip # which GLS trip is selecting

LoggerModel<-info$Modelo #logger model
Producer<- info$Producer #company

# Colony infomation
col.name <- info$Colony
col_lat <- info$Colony.lat
col_lon <- info$Colony.lon

# Dates information
year_rec <- info$Year_rec # which year we recovered the GLS

start_date<-as.Date(info$start_date, tz = 'GMT') # date start the trip
start_year<-year(start_date) # year start the trip
end_date <-as.Date(info$end_date, tz = 'GMT') # date end the trip
end_year<-year(end_date) # year end the trip

cat(paste0('Processing geo (',i,') ', geo_trip, ' for year ',
           year_rec, ' from ', col.name),  '\n')

#### 2.1 Load .trn file ####
trn<-read_csv(trnfile_c, col_names = T, show_col_types = F)
twl<- trn %>%
  filter(Twilight >= start_date & Twilight <= end_date)%>% # cut data
  distinct(Twilight, .keep_all = TRUE)%>%  # remove duplicated rows 
  arrange(Twilight)

# If any GLS has not calibration parameters, we select the mean for this producer, colony and year
calib.param <- select_sunAngle(producer = Producer, col.name, year = year_rec,
                                calib.sum, calibs, param = T)
zenith <- calib.param$zenith_mean
zenith0 <- calib.param$zenith0_mean
alpha <- calib.param$alpha
```

### 4.2. Determine tol parameter

As Lisosvki et al. (2019) mentioned, tol is a parameter to modulate the tolerance on the sine of the solar declination

```{r}

# Select equinox peridos

Sep_eq <- as.Date(paste0(start_year, '-09-21'))
Mar_eq <- as.Date(paste0(end_year, '-03-21'))

tol_res <- calc_tol(twl, n = 0.001, d = 20,
                    Sep_eq, Mar_eq, k1 = 0.95, k2= 0.5, 
                    plot = T)

tol_res$plot
```

### 4.3. SGAT initial path

```{r}

# Dividning trip to calculate path with different tol values for each equinox
if(!is.null(tol_res$tol_mar)){
  twl_1 <- twl %>%
    filter(Twilight >= start_date & Twilight <= paste0(end_year,'-01-01'))
  twl_2 <- twl %>%
    filter(Twilight > paste0(end_year,'-01-01') & Twilight <= end_date)
  
  path_1 <- thresholdPath(twl_1$Twilight, twl_1$Rise, zenith = zenith, tol=tol_res$tol_sep)
  x0_1 <- path_1$x
  t1 <- path_1$time
    
  path_2 <- thresholdPath(twl_2$Twilight, twl_2$Rise, zenith = zenith, tol=tol_res$tol_mar)
  x0_2 <- path_2$x
  t2 <- path_2$time
  
  x0 <- rbind(x0_1, x0_2)
  time <- c(t1, t2)
  
} else {
  path <- thresholdPath(twl$Twilight, twl$Rise, zenith = zenith, tol=tol_res$tol_sep)
  x0 <- path$x
  time <- path$time
}


cat('Initial path determined','\n')

# Plot the track
p1 <- plot_SGAT(x0, time, d = 20, Sep_eq, Mar_eq, world, col_lat, col_lon)
p1
```

### 4.4. Save SGAT raw track

```{r}
raw.points<-cbind(time, x0)

colnames(raw.points)<-c('Date_Time', 'lon', 'lat')

if(is.null(raw.points)){
  message(paste0('There is no trip from ', geoID, ' - Skipped \n'))
  mdata2[i, 24] <- 'No'
  next
}else{
  mdata2[i, 24] <- 'Yes'
  cat(paste0('Raw points from ', geoID, ' plotted \n'))
}

# Add information
raw.points<-bind_cols(Ring = ring, Species = sp, Colony = col.name, geoID = geoID,
                      geo_trip = geo_trip, Producer = Producer,
                      start_year = start_year, end_year = end_year, 
                      tol_sep = tol_res$tol_sep, tol_mar = tol_res$tol_mar, raw.points)
```

### 4.5. Clean SGAT positions through colony days information

```{r}
#### Define known locations ####
# Here we assign that the bird is in the colony the days that is not recorded light (inside the nest)
# and during the breeding period the dry periods of the day were longer than 12 hours.

cdays <- col.days%>%
  dplyr::filter(geo_trip == .env$geo_trip)%>%
  filter(light == T | wetdry == T)


# Create an auxiliary dataframe, to filter the dates 
auxi <- data.frame(date = as.Date(time))
auxi <-cbind(auxi, as.data.frame(x0),row = row_number(auxi))
auxi <- auxi %>%
  filter(date %in% cdays$date)

auxi <- left_join(auxi, cdays[,c(3,6,9)])
# Define the locations in the colony as TRUE
fixedx <- rep(0, nrow(x0))

fixedx[auxi[is.na(auxi$wetdry),]$row] <- 3
fixedx[auxi[auxi$wetdry == T,]$row] <- 2
fixedx[auxi[auxi$light == T,]$row]  <- 1
fixedx[1:2] <- 1 # first two location estimates

# The last line assign the last position in the colony
# If we use resuscitated GLSs we do not use
if(end_date == as.Date(twl$Twilight[nrow(twl)]))
{fixedx[(nrow(x0) - 1):nrow(x0)] <- 1 # last two location estimates
}else{
  fixedx[(nrow(x0) - 1):nrow(x0)] <- 0 # be sure resuscitated files have not return to the colony
}

# Generate a random point 50 km away from the given point
for (k in which(fixedx != 0)) {
  code <- fixedx[k]
  if(code == 1) {
    x0[k, 1] <- col_lon
    x0[k, 2] <- col_lat
  } else if (code == 2){
    random_p <- geosphere::destPoint(p = c(col_lon, col_lat),
                                     b = runif(1, 0, 360), d = 100*1000)
    lat <- as.numeric(random_p[2])
    lon <- as.numeric(random_p[1])
    
    x0[k, 1] <- lon
    x0[k, 2] <- lat
  } 
}
```

```{r, echo=TRUE, include=TRUE, eval=FALSE}
which(fixedx == TRUE)
x0[fixedx, 1] <- col_lon
which(x0 == col_lon)
x0[fixedx, 2] <- col_lat
which(x0 == col_lat)
```

```{r}

# Plot the track
p2 <- plot_SGAT(x0, time, d = 20, Sep_eq, Mar_eq, world, col_lat, col_lon)
p2
cat('Colony days corrected','\n')

# Raw trip
raw.points2<-cbind(time, x0)
colnames(raw.points2)<-c('Date_Time', 'lon', 'lat')

# Add information
raw.points2<-bind_cols(Ring = ring, Species = sp, Colony = col.name, geoID = geoID,
                       geo_trip = geo_trip, Producer = Producer,
                       start_year = start_year, end_year = end_year, 
                       tol_sep = tol_res$tol_sep, tol_mar = tol_res$tol_mar,
                       raw.points2)

```

## 5. Save SGAT raw tracks


```{r, echo=TRUE, include=TRUE, eval=FALSE}
raw_trips.list[i]<-list(raw.points)
raw_trips2.list[i]<-list(raw.points2)

raw_trips <- plyr::ldply(raw_trips.list, data.frame)
raw_trips2 <- plyr::ldply(raw_trips2.list, data.frame)

write_csv(raw_trips, '/output/Raw_migratory_trips_tol_v3.csv')
write_csv(raw_trips2, '/output/Raw_migratory_trips_tol_cdays_v3.csv')
```
