---
title: "Automated migratory classification"
author: "Diego Vicente-Sastre"
format: html
editor: visual
date-modified: today
# bibliography: references.bib
execute: 
  echo: true
  warning: false
  message: false
---

## 

## 1. Introduction: segmentation and clustering migration behaviours

## 2. Load packages and functions

Every time we do not need to install and load packages from GitHub or other external repository, we use `pacman` function. For the following code, we need two main packages `adehabitatHR` and `segclust2d` packages

```{r}
pacman::p_load('devtools',
               'tidyverse', 'plyr', 'lubridate', 'readxl', 'readr', # data carpenter
                'sf', 'spData', # spatial data
               'adehabitatHR',  # migrate behaviour and kernels
               'segclust2d', # spatial segmentation and clustering
               'units', 
               'ggplot2', 'ggpubr', 'ggspatial',# plots
               'patchwork', 'cowplot', # composing plots
               'caTools'# runmean and runmin calcs
)
```

We also load own functions, specifically designed for working with GLS data.

```{r}
source('functions/02_FUNCTIONS_automated_phenology.R') 
```

## 3. Load tracking data and metada

```{r}
mdata3 <- read_csv('Raw Data/GLS.files3.csv', col_names = TRUE)

# dataset <- 'v1'
tracks <- read_csv('output/Raw_SGAT_tracks_tol_cdays.csv', col_names = TRUE)

```


## 4. Defining trajectory metrics
We define the trajectory metrics that we will use to classify the migratory behaviour of the individuals.
```{r}
# Core variables (non-optional)
core.v <- c("straightness", "sinuosity", "ang_vector")

# Define the non-paired variables
add.v <- c("mu.dist", "sd.dist", "Emax", "sq_disp", "R2n_sd", "dc_mean", "dc_sd")

# Paired variables
pair.v <- list(
  mu_xy = c("mu.x", "mu.y"),
  sd_xy = c("sd.x", "sd.y"),
  mu_dxy = c("mu.dx", "mu.dy"),
  sd_dxy = c("sd.dx", "sd.dy")
)

# Create a list of all variable groups (paired and non-paired)
all_variable_groups <- c(names(pair.v), add.v)
```

```{r}
i= 1


geo_trip <- ids[i] #GLS id

info<- GLS.files%>%dplyr::filter(geo_trip == .env$geo_trip)
geoID <- info$geo_alfanum
Ring<- info$Ring #Bird ID
Sp <- info$Sp
resucitat<-info$Resucitat

# Colony info
col.name <- info$Colony
col.lon = info$Longitude
col.lat = info$Latitude

colony_sf<-colony%>%
  st_as_sf(coords = c("Longitude", "Latitude"),
           crs = projections$WGS84, agr = "constant")

cdays <- colony_days%>%
  dplyr::filter(geo_trip == .env$geo_trip & light == T)

trip<-data_clean%>%
  dplyr::filter(ID == geo_trip)

#Ensure first position is the colony
# trip$Longitude[1] <- colony$Longitude
# trip$Latitude[1] <- colony$Latitude
if(nrow(trip) == 0){
  message('There is no trip positions for geo (', i, ') ', geo_trip, '\n')
  next
}

start_year <-  year(trip$DateTime[1]) #Trip start
end_year <- year(trip$DateTime[nrow(trip)]) #Trip end

cat('Processing geo (', i,') ', geo_trip, 'since', start_year, 'to', end_year, 
    ' from ', col.name,  '\n')
# 6. Segmentation/Clustering of the track ####

# Trajectory format 
# R2n	= the squared net displacement between the current relocation 
# and the first relocation of the trajectory
ltraj <- as.ltraj(xy = trip[,c('Longitude','Latitude')],
                  date = trip$DateTime,
                  id = trip$ID)
df_trip <- ltraj[[1]]
if(geo_trip == '1600004001') df_trip <- df_trip[-c(1,2),]




# Determine if there are movement during equinox days
# Select equinoxes
Sept_eq <- pull(equinox[equinox$Year == start_year,2])
Mar_eq <- pull(equinox[equinox$Year == end_year,4])


df_trip2 <- clean_positions(data = df_trip, dist_threshold = 40)

# if(geo_trip %in% c('1606001001', '19204002001', '2293003001',
#                    '23287001001','23287001001', '23290004001',
#                    '23349001001', 'BD811003001', 'BH072001001', 
#                    'BH176002001', 'N814001001', 'X241001001')) next # add to bad trips or check if they need a different tol

if(geo_trip == 'BL224002001') {
  df_trip <- df_trip %>%
    filter(dy != 0)
  ltraj <- as.ltraj(xy = df_trip[,c('x','y')],
                    date = df_trip$date,
                    id = geo_trip)
  df_trip <- ltraj[[1]]
  df_trip2 <- df_trip %>%
    filter(dist < 30)
  
}
# Subset to use after seg/clust analysis
df_cdays <-  df_trip2%>%
  filter(as.Date(date) %in% cdays$date)%>%
  add_column(Breed = 'B')

df_trip3 <- remove_duplicates(data = df_trip2, equinox_days = 20, 
                              cdays = cdays,
                              Sept_eq = Sept_eq, Mar_eq = Mar_eq, dup_threshold = 4)
if(any(duplicated(tail(df_trip3$y, 5)))){
  df_trip3 <- df_trip3[1:(nrow(df_trip3) - 5),]
} else if (any(duplicated(head(df_trip3$y, 5)))){
  df_trip3 <- df_trip3[5:nrow(df_trip3),]
}

# Correct R2n values, according to the real distance to the colony
df_trip3 <- df_trip3 %>%
  mutate(R2n = (x - col.lon)^2 + (y - col.lat)^2)

# if(geo_trip == 'BM283002001') df_trip3 <- df_trip3 %>%filter(dist < 10)
if(geo_trip == 'BD137001001') df_trip3 <- df_trip3 %>%filter(dist < 25)
if(geo_trip == 'X248001001') df_trip3 <- df_trip3[-c(1,2),]
if(geo_trip == '19094002001') df_trip3 <- df_trip3[1:521,]
if(geo_trip == '1602002001') df_trip3 <- df_trip3[36:(nrow(df_trip3)-6),]
# if(geo_trip == 'BM271004001') df_trip3 <- df_trip3[1:(nrow(df_trip3)-1),]

# Segmentation/clustering of the hole trip
mode_segclust <- segclust(df_trip3,
                          Kmax = 15,
                          lmin = 5,
                          ncluster = 5:10,
                          seg.var = c('x', 'y'),
                          # seg.var = c('R2n','dist'),
                          order.var = 'x',
                          diag.var = c('dx', 'dy', 'R2n', 'x', 'y', 'dist'),
                          subsample = F)

# Results of segmentation/clustering
# Plots
segclust_plot <- plot(mode_segclust) # variables distribution
map_seg <- segmap(mode_segclust, pointsize = 3, 
                  coord.names = c("x","y"), interactive = F
) # map segmentation
# segclust_plot
# map_seg
# Check likelihood of alternative number of segment possible. There should
# be a clear break if the segmentation is good
bic <- BIC(mode_segclust)
bic_max <- bic %>% filter(BIC == max(BIC))
BIC_plot<-plot_BIC(mode_segclust)

# Number of segments
nseg <- bic_max$nseg
# Number of clusters
ncluster <- bic_max$ncluster

# 7. Re-classification of the segments ####

# Segments' reclassification to infer biological meaning
# segs <- segment(mode_segclust)
segments1 <- segment_classification(mode_segclust = mode_segclust,
                                    # projection = projection,
                                    option.by = 'seg',
                                    col.lon = col.lon,
                                    col.lat = col.lat)
# Equinox correction
segments2 <- equinox_interference(df_seg = segments1$df_clust, track = df_trip3,
                                  Sept_eq, Mar_eq)


segments3 <- segment_correction(df_seg = segments2)

segments3 <- segments3 %>%
  mutate(
    Check = case_when(
      Confidence < 2 ~ 'Check',
      Confidence >= 2 & Eq & !(row_number() %in% c(1, n())) & 
        (lag(Breed) != Breed | lead(Breed) != Breed) ~ 'Check',
      TRUE ~ 'Ok'
    )
  )
segments3 <- cbind(ID = geo_trip, Sp = Sp, segments3)

seg.list2[i] <- list(segments3)

# Resulting data.frame
df_trip4 <-segclust2d::augment(mode_segclust)

df_trip5 <- df_trip4 %>%
  # left_join(., df_trip4)%>%
  # Join Breed to hole track
  mutate(
    Breed = pmap_chr(
      list(date = date),
      ~ {
        idx <- which(segments3$f_day <= ..1 & segments3$l_day >= ..1)
        if (length(idx) > 0) {
          segments3$Breed[idx[1]]
        } else {
          NA_character_
        }
      }
    )) %>%
  distinct() %>%
  # arrange(date) %>%
  add_column(ID = geo_trip) %>%
  dplyr::select(ID, everything()) %>%
  # tidyr::fill(Breed, .direction = "down")%>%
  # tidyr::fill(state, .direction = "down")%>%
  # tidyr::fill(state_ordered, .direction = "down")%>%
  mutate(Breed2 = case_when(
    Breed == 'mig' & cumsum(replace_na(Breed == 'W', FALSE)) > 0 & 
      rev(cumsum(rev(replace_na(Breed == "W", FALSE)))) == 0 ~ 'mig2',
    Breed == "mig" & rev(cumsum(rev(replace_na(Breed == "W", FALSE)))) > 0 ~ "mig1", 
    TRUE ~ Breed
  )) %>%
  mutate(Equinox = ifelse(
    date >= (as.Date(Sept_eq) - days(20)) & 
      date <= (as.Date(Sept_eq) + days(20)), 'Sep',
    ifelse(
      date >= (as.Date(Mar_eq) - days(20)) & 
        date <= (as.Date(Mar_eq) + days(20)), 'Mar', 'No'
    )
  ))

tracks_list[i] <- list(df_trip5)

plot_list <- plot_segmentation(segclust = segclust_plot, map = map_seg, 
                               segclass = segments3, 
                               col.lon = colony$Longitude, col.lat = colony$Latitude,
                               breed = T, data = df_trip5,
                               world = world, option = 'all')


# plot_save <- plot_list$segclust | (plot_list$map / plot_list$segclass) | plot_list$breed
plot_save <- plot_list$map | plot_list$breed


ggsave(plot = plot_save,
       filename = paste0(wd,'/Plots/Phenology/', 
                         geo_trip, '_', end_year,'_', Sp, '_', col.name, '_',
                         'seg-clust.jpg'),
       width = 42, height = 18, units = 'cm')

cat('First plot saved \n')

# Detect resident individuals
if(all(segments3$Breed != 'W') | all(segments3$Breed != 'B') |
   all(segments3$index_seg2[which(segments3$Breed == 'W')] < 20)){
  Migrant <- 'Resident'
  
  phen <- data.frame(Ring = Ring, Sp = Sp, Colony = col.name,
                     ID = geo_trip, Migrant = Migrant,  
                     BA_dep = as.POSIXct(NA), WA_arr = as.POSIXct(NA),
                     WA_dep = as.POSIXct(NA), BA_arr = as.POSIXct(NA)
  )
  phen <- phen %>%
    mutate(across(c(BA_dep, WA_arr, WA_dep, BA_arr), as.Date))
  
  phen.list[i] <- list(phen)
  
  cat('Phenology from geo (', i, ')', geo_trip, 'cannot be calculated \n')
  cat('Track classified as resident \n')
} else { 
  Migrant <- 'Migrant'
  cat('Track classified as migrant \n')
}

# Summary of segmentation/clustering information
seg_df <- data.frame(Ring = Ring, Sp = Sp, Colony = col.name,
                     ID = geo_trip, Migrant = Migrant,
                     n_seg = nseg, n_cluster = ncluster, 
                     Breed_class = paste(segments3$Breed, collapse = '-'),
                     n_seg_B = length(segments3$Breed[segments3$Breed == 'B']),
                     n_seg_mig = length(segments3$Breed[segments3$Breed == 'mig']),
                     n_seg_W = length(segments3$Breed[segments3$Breed == 'W']),
                     Confidence_0 = mean(segments2$Confidence, na.rm = T)/3,
                     Confidence = mean(segments3$Confidence, na.rm = T)/3,
                     Metrics = paste(segments1$metrics, collapse = '-')
                     
)
seg.list[i] <- list(seg_df)

if(Migrant == 'Resident') next

pBA_dep <- first(which(df_trip5$Breed == 'mig'))
pWA_arr <- first(which(df_trip5$Breed == 'W'))
pWA_dep <- last(which(df_trip5$Breed == 'W')) + 1
pBA_arr <- last(which(df_trip5$Breed == 'mig')) + 1

BA_dep <- as.Date(df_trip5$date[pBA_dep])
WA_arr <- as.Date(df_trip5$date[pWA_arr])
WA_dep <- as.Date(df_trip5$date[pWA_dep])
BA_arr <- as.Date(df_trip5$date[pBA_arr])

phen <- data.frame(Ring = Ring, Sp = Sp, Colony = col.name,
                   ID = geo_trip, Migrant = Migrant,  
                   BA_dep = BA_dep, WA_arr = WA_arr,
                   WA_dep = WA_dep, BA_arr = BA_arr
)
phen <- phen %>%
  mutate(across(c(BA_dep, WA_arr, WA_dep, BA_arr), as.Date))
```

