---
title: "Automated migratory classification"
author: "Diego Vicente-Sastre"
format: html
editor: visual
date-modified: today
# bibliography: references.bib
execute: 
  echo: true
  warning: false
  message: false
---

## 

## 1. Introduction: segmentation and clustering migration behaviours

## 2. Load packages and functions

Every time we do not need to install and load packages from GitHub or other external repository, we use `pacman` function. For the following code, we need two main packages `adehabitatHR` and `segclust2d` packages

```{r}
pacman::p_load('devtools',
               'tidyverse', 'plyr', 'lubridate', 'readxl', 'readr', # data carpenter
                'sf', 'spData', # spatial data
               'adehabitatHR',  # migrate behaviour and kernels
               'segclust2d', # spatial segmentation and clustering
               'units', 
               'ggplot2', 'ggpubr', 'ggspatial',# plots
               'patchwork', 'cowplot', # composing plots
               'caTools'# runmean and runmin calcs
)
```

We also load own functions, specifically designed for working with GLS data.

```{r}
source('functions/02_FUNCTIONS_automated_phenology.R') 
```

## 3. Read metadata and calibration parameters
