---
title: "Automated migratory classification"
author: "Diego Vicente-Sastre"
format: html
editor: visual
date-modified: today
# bibliography: references.bib
execute: 
  echo: true
  warning: false
  message: false
---

## 

## 1. Introduction: segmentation and clustering migration behaviours

## 2. Load packages and functions

Every time we do not need to install and load packages from GitHub or other external repository, we use `pacman` function. For the following code, we need two main packages `adehabitatHR` and `segclust2d` packages

```{r}
pacman::p_load('devtools',
               'tidyverse', 'plyr', 'lubridate', 'readxl', 'readr', # data carpenter
                'sf', 'spData', # spatial data
               'adehabitatHR',  # migrate behaviour and kernels
               'segclust2d', # spatial segmentation and clustering
               'units', 
               'ggplot2', 'ggpubr', 'ggspatial',# plots
               'patchwork', 'cowplot', # composing plots
               'caTools'# runmean and runmin calcs
)
```

We also load own functions, specifically designed for working with GLS data.

```{r}
source('functions/02_FUNCTIONS_automated_phenology.R') 
```

## 3. Load tracking data and metada

```{r}
mdata3 <- read_csv('input/GLS.files3.csv', col_names = TRUE)

# dataset <- 'v1'
tracks <- read_csv('output/Raw_SGAT_tracks_tol_cdays.csv', col_names = TRUE)

```

## 4. Defining trajectory metrics

We define the trajectory metrics that we will use to classify the migratory behaviour of the individuals. Algorithm use straightness, sinuosity and vector angle as defect, and it is suported in other accesorial variables that can refine the classification.

```{r}

# Define the non-paired metrics
add.m <- c("mu.dist", "sd.dist", "Emax", "sq_disp", "R2n_sd", "dc_mean", "dc_sd")

# Paired metrics
pair.m <- list(
  mu_xy = c("mu.x", "mu.y"),
  sd_xy = c("sd.x", "sd.y"),
  mu_dxy = c("mu.dx", "mu.dy"),
  sd_dxy = c("sd.dx", "sd.dy")
)

# Create a list of all metrics groups (paired and non-paired)
add.metrics <- c(names(pair.m), add.m)
```

## 4. Classification of tracking data

We will work again through a loop.

### 4.1. Read track information

```{r}
i= 1

info<- mdata3[i,]

ring<- info$Ring # individual
sp <- info$Sp # species
geoID<-info$geo_alfanum # Which GLS is processing
geo_trip<-info$geo_trip # which GLS trip is selecting

# Colony info
col.name <- info$Colony
col.lon = info$Colony.lon
col.lat = info$Colony.lat

track <- tracks%>%
  filter(geo_trip == .env$geo_trip)

start_year <-  year(track$Date_Time[1]) #Trip start
end_year <- year(track$Date_Time[nrow(track)]) #Trip end

cat('Processing geo (', i,') ', geo_trip, 'since', start_year, 'to', end_year, 
    ' from ', col.name,  '\n')
```

### 4.2. Prepare track to be segmented

Segmentation and clustering by segclust2d package requires:

-   **Trajectory class** object

-   **Avoid low variance** segments

Additionally, if we are managing GLS data, we have to ensure **first position is at the colony**. In that way, we use Net Squared Colony Distance (NSCD) instead of Net Squared Distance (NSD).

##### Convert to trajectory class

```{r}
# Trajectory format 
# R2n	= the squared net displacement between the current relocation and the first relocation of the trajectory (NSD)
ltraj <- as.ltraj(xy = as.matrix(track[,c('lon','lat')]),
                  date = track$Date_Time,
                  id = track$geo_trip)
track_df <- ltraj[[1]]
```

Transforming into ltraj object, we can get two things at once: (1) x and y columns as coordinates of the track and (b) several movement metrics can be useful for segmentation, clustering and classification of the segments.

#### Cleaning tracking data for segmentation and clustering
```{r}
# Determine if there are movement during equinox days
# Select equinoxes
Sep_eq <- as.Date(paste0(start_year, '-09-21'))
Mar_eq <- as.Date(paste0(end_year, '-03-21'))
```

Optionally we can remove colony days if they cause low-variance segments that `segclust` function cannot handle.

```{r, eval= FALSE, include= FALSE}
# cdays <- colony_days%>%
#   dplyr::filter(geo_trip == .env$geo_trip)
```

First, we remove erroneus positions, where the distance among them is too large. Then, we remove possible pseudo-duplicates, low-variance segments, that can be caused by the modelling of positions along the equinox periods, the correction of the positions along colony days or another reason that cause > 4 consecutive positions with the same coordinates.

```{r}
# Clean erroneus positions
track_df2 <- clean_positions(data = track_df, dist_threshold = 40)

track_df3 <- remove_duplicates(data = track_df2, equinox_days = 20, 
                              # cdays = cdays,
                              Sep_eq = Sep_eq, Mar_eq = Mar_eq, dup_threshold = 4)
```

### 4.3. Segmentation and clustering

```{r}

mode_segclust <- segclust(track_df3,
                          Kmax = 15,
                          lmin = 5,
                          ncluster = 5:10,
                          seg.var = c('x', 'y'),
                          order.var = 'x',
                          diag.var = c('dx', 'dy', 'R2n', 'x', 'y', 'dist'),
                          subsample = F)

# Results of segmentation/clustering

# Plots
segclust_plot <- plot(mode_segclust) # variables distribution
map_seg <- segmap(mode_segclust, pointsize = 3, 
                  coord.names = c("x","y"), interactive = F) 

bic_plot <- plot_BIC(mode_segclust)
segclust_plot + (map_seg / bic_plot)

```

```{r}
# 7. Re-classification of the segments ####

# Segments' reclassification to infer biological meaning
# segs <- segment(mode_segclust)
segments1 <- segment_class(mode_segclust = mode_segclust,
                           option.by = 'seg',
                           add.metrics = add.metrics,
                           col.lon = col.lon,
                           col.lat = col.lat)
# Equinox correction
segments2 <- equinox_interference(df_seg = segments1$df_clust, 
                                  track = track_df3,
                                  Sep_eq, Mar_eq)


segments3 <- segment_correction(df_seg = segments2)

segments3 <- segments3 %>%
  mutate(
    Check = case_when(
      Confidence < 2 ~ 'Check',
      Confidence >= 2 & Eq & !(row_number() %in% c(1, n())) & 
        (lag(Breed) != Breed | lead(Breed) != Breed) ~ 'Check',
      TRUE ~ 'Ok'
    )
  )
segments3 <- cbind(ID = geo_trip, Sp = sp, segments3)
```


```{r}
seg.list2[i] <- list(segments3)
```


```{r}
# Resulting data.frame
df_trip4 <-segclust2d::augment(mode_segclust)

df_trip5 <- df_trip4 %>%
  # left_join(., df_trip4)%>%
  # Join Breed to hole track
  mutate(
    Breed = pmap_chr(
      list(date = date),
      ~ {
        idx <- which(segments3$f_day <= ..1 & segments3$l_day >= ..1)
        if (length(idx) > 0) {
          segments3$Breed[idx[1]]
        } else {
          NA_character_
        }
      }
    )) %>%
  distinct() %>%
  # arrange(date) %>%
  add_column(ID = geo_trip) %>%
  dplyr::select(ID, everything()) %>%
  # tidyr::fill(Breed, .direction = "down")%>%
  # tidyr::fill(state, .direction = "down")%>%
  # tidyr::fill(state_ordered, .direction = "down")%>%
  mutate(Breed2 = case_when(
    Breed == 'mig' & cumsum(replace_na(Breed == 'W', FALSE)) > 0 & 
      rev(cumsum(rev(replace_na(Breed == "W", FALSE)))) == 0 ~ 'mig2',
    Breed == "mig" & rev(cumsum(rev(replace_na(Breed == "W", FALSE)))) > 0 ~ "mig1", 
    TRUE ~ Breed
  )) %>%
  mutate(Equinox = ifelse(
    date >= (as.Date(Sept_eq) - days(20)) & 
      date <= (as.Date(Sept_eq) + days(20)), 'Sep',
    ifelse(
      date >= (as.Date(Mar_eq) - days(20)) & 
        date <= (as.Date(Mar_eq) + days(20)), 'Mar', 'No'
    )
  ))

tracks_list[i] <- list(df_trip5)

plot_list <- plot_segmentation(segclust = segclust_plot, map = map_seg, 
                               segclass = segments3, 
                               col.lon = colony$Longitude, col.lat = colony$Latitude,
                               breed = T, data = df_trip5,
                               world = world, option = 'all')


# plot_save <- plot_list$segclust | (plot_list$map / plot_list$segclass) | plot_list$breed
plot_save <- plot_list$map | plot_list$breed


ggsave(plot = plot_save,
       filename = paste0(wd,'/Plots/Phenology/', 
                         geo_trip, '_', end_year,'_', Sp, '_', col.name, '_',
                         'seg-clust.jpg'),
       width = 42, height = 18, units = 'cm')

cat('First plot saved \n')

# Detect resident individuals
if(all(segments3$Breed != 'W') | all(segments3$Breed != 'B') |
   all(segments3$index_seg2[which(segments3$Breed == 'W')] < 20)){
  Migrant <- 'Resident'
  
  phen <- data.frame(Ring = Ring, Sp = Sp, Colony = col.name,
                     ID = geo_trip, Migrant = Migrant,  
                     BA_dep = as.POSIXct(NA), WA_arr = as.POSIXct(NA),
                     WA_dep = as.POSIXct(NA), BA_arr = as.POSIXct(NA)
  )
  phen <- phen %>%
    mutate(across(c(BA_dep, WA_arr, WA_dep, BA_arr), as.Date))
  
  phen.list[i] <- list(phen)
  
  cat('Phenology from geo (', i, ')', geo_trip, 'cannot be calculated \n')
  cat('Track classified as resident \n')
} else { 
  Migrant <- 'Migrant'
  cat('Track classified as migrant \n')
}

# Summary of segmentation/clustering information
seg_df <- data.frame(Ring = Ring, Sp = Sp, Colony = col.name,
                     ID = geo_trip, Migrant = Migrant,
                     n_seg = nseg, n_cluster = ncluster, 
                     Breed_class = paste(segments3$Breed, collapse = '-'),
                     n_seg_B = length(segments3$Breed[segments3$Breed == 'B']),
                     n_seg_mig = length(segments3$Breed[segments3$Breed == 'mig']),
                     n_seg_W = length(segments3$Breed[segments3$Breed == 'W']),
                     Confidence_0 = mean(segments2$Confidence, na.rm = T)/3,
                     Confidence = mean(segments3$Confidence, na.rm = T)/3,
                     Metrics = paste(segments1$metrics, collapse = '-')
                     
)
seg.list[i] <- list(seg_df)

if(Migrant == 'Resident') next

pBA_dep <- first(which(df_trip5$Breed == 'mig'))
pWA_arr <- first(which(df_trip5$Breed == 'W'))
pWA_dep <- last(which(df_trip5$Breed == 'W')) + 1
pBA_arr <- last(which(df_trip5$Breed == 'mig')) + 1

BA_dep <- as.Date(df_trip5$date[pBA_dep])
WA_arr <- as.Date(df_trip5$date[pWA_arr])
WA_dep <- as.Date(df_trip5$date[pWA_dep])
BA_arr <- as.Date(df_trip5$date[pBA_arr])

phen <- data.frame(Ring = Ring, Sp = Sp, Colony = col.name,
                   ID = geo_trip, Migrant = Migrant,  
                   BA_dep = BA_dep, WA_arr = WA_arr,
                   WA_dep = WA_dep, BA_arr = BA_arr
)
phen <- phen %>%
  mutate(across(c(BA_dep, WA_arr, WA_dep, BA_arr), as.Date))
```
