---
title: "Phenology dates of a migratory animal track"
author: "Diego Vicente-Sastre"
format: html
editor: visual
date-modified: today
# bibliography: references.bib
execute: 
  echo: true
  warning: false
  message: false
---

## 1. Introduction: Kernel Density Estimation

## 2. Load packages and functions

```{r}

pacman::p_load('tidyverse', 'dplyr', 'lubridate', 'readxl', 'readr', # data carpenter
               'sf', # spatial data
               'adehabitatHR', 'track2KBA', # migrate behaviour and kernels
               'ggplot2', 'ggpubr', 'ggspatial',# plots
               'patchwork', 'cowplot' # composing plots
)
```

We also load own functions designed to detect key phenological dates of migratory animal tracks

```{r}
source('functions/04_FUNCTIONS_Phenology_dates.R')
```

## 3. Load tracking data and metadata

Besides metadata and tracks, we load the **wintering areas** shapefile. This file contains the different ecoregions of the Atlantic ocean where seabirds like shearwaters can winter. These ecoregions were defined by ...  and the shapefile is adapted by Virginia et al. (2025), where she included two ecoregions more (Equatorial Atlantic and South Atlantic). These ecoregions can be adapted for other species and in the context of other oceans and seas.

```{r}
#Metadata
mdata3 <- read_csv('input/GLS.files3.csv', col_names = TRUE)

# Tracking data
tracks.ph <- read_csv('output/Phenology_tracks_corrected.csv')
ids <-unique(tracks.ph$ID)

# Wintering areas and ecoregions
w.areas <- read_sf(dsn = "input/wintering_areas")
ecoregions <- w.areas %>% dplyr::select(ECOREGION)

world <- rnaturalearth::ne_countries(scale = "large", returnclass = "sf")
sf_use_s2(FALSE)
```


## 4. Migration phenology estimation

### 4.1. Load track data and metadata
```{r}
i = 1

info<- mdata3 %>% filter(geo_trip == ids[i])

geoID <- info$geo_alfanum
geo_trip <- info$geo_trip
ring<- info$Ring #Bird ID
sp <- info$Sp

cat('Processing geo ', geo_trip, '(', i, ') \n')

track.ph <- tracks.ph %>% filter(ID == ids[i])

start_year <- year(track.ph$date[1])
end_year <- year(track.ph$date[nrow(track.ph)]) 


# Colony info
col.name <- info$Colony
col.lat <- info$Colony.lat
col.lon <- info$Colony.lon
```

### 4.2. Check and assing phenological periods

In this step we count the number of periods of each phenological behaviour. Ideally, a migratory track will contain a pre-migration breeding period (B1), a post-nuptial migration period (mig1), a wintering period (W1), a pre-nuptial migration period (mig2) and a post-migratio breeding period (B2). However, in some cases, the track can contain more than one wintering period (e.g.,  the W1, W2, W3), and therefore, more than one migration period between wintering periods. These migration periods between wintering areas we still consider as post-nuptial migration periods (mig1).

It is very important to get the results that this part of the procedure is going well. Sometimes there are little mistakes in th assignation of the phenological stage to some positions (many times as a consequence of the supervised re-classification). For example, if there are four positions assigned to "mig1" inside the "B1" period, we will get two breeding periods: "B1", "mig1", "B2" and then, the real "B2" period will be "B3".

Because all of this, we check and correct this mistakes and then count the number of each phenological stages.


```{r}
# Correct possible mistakes

track.ph2 <- check_phen(track.ph, "Breed2")

track.ph2$Breed3 <- phen_periods(track.ph2$Breed2)

# Identify unique B and W stages
BW_periods <- unique(track.ph2$Breed3[grep("^B|^W", track.ph2$Breed3)])
B_periods <- grep("^B", BW_periods, value = TRUE)
W_periods <- grep("^W", BW_periods, value = TRUE)

if(B_periods[1] == 'B1' & track.ph2$Breed2[1] != 'B'){
  Bperiods <- 'B2'
  track.ph2 <- track.ph2 %>%
    mutate(Breed3 = case_when(Breed3 == 'B1' ~ 'B2',
                              T ~ Breed3))
}
if(length(W_periods) == 0){
  Wperiods <- NULL
}

BW_periods
```

### 4.3. Data format

We are going to use `track2KBA` package to calculate the KDE. This package needs the data in a specific format. We will use the function `formatFields` to format the tracking data. 

```{r}
track_f <- formatFields(dataGroup = track.ph2, 
                          fieldID   = "ID", 
                          fieldDateTime = "date", 
                          fieldLon  = "x", 
                          fieldLat  = "y" )

```

Optionally, we can plot the track with the phenology classification of the points. We can save this plot, but it is useful to see the track classified before KDE calculation.

```{r}
####3. Phenology plot 1  #####
breed_df <- track.ph2 %>%
  mutate(
    Shape = case_when(Equinox == 'Sep' ~ 24,
                      Equinox == 'Mar' ~ 25,
                      Equinox == 'No' ~ 21
    ),
    Color = case_when(Breed2 == 'B' ~ 'orange2',
                      Breed2 == 'mig1' ~ 'grey40',
                      Breed2 == 'mig2' ~ 'grey70',
                      Breed2 == 'W' ~ 'steelblue',
    ),
    lab = case_when(Breed2 == 'B' ~ 'Breeding',
                    Breed2 == 'mig1' ~ 'Migration post',
                    Breed2 == 'mig2' ~ 'Migration pre',
                    Breed2 == 'W' ~ 'Wintering',
    )
  )
breaks <- unique(breed_df$Color)
labels <- unique(breed_df$lab)

# Breeding vs non-breeding plot
breed_plot <- ggplot()+
  # World
  {if(!is.null(world))
    geom_sf(data = world, fill=grey(0.9), color=grey(0.6), lwd = 0.2)
  } +
  
  coord_sf(xlim = c(min(breed_df$x), max(breed_df$x)), 
           ylim = c(min(breed_df$y), max(breed_df$y)), expand = TRUE) +
  # Trip
  geom_path(data= breed_df, aes(x, y))+
  {if(length(unique(breed_df$Shape)) > 1)
    geom_point(data= breed_df, aes(x, y, fill = Color,
                                   shape = Shape), size = 3)
  } +
  {if(length(unique(breed_df$Shape)) == 1)
    geom_point(data= breed_df, aes(x, y, fill = Color),
               shape = 21, size = 3)
  }+
  # Colors
  scale_fill_identity(guide = 'legend',
                      name = 'Track classification',
                      breaks = breaks,
                      labels = labels
  )+
  # Shapes
  scale_shape_identity(breaks = c(21, 25, 22),
                       labels = c('No', 'March', 'September'),
                       name = 'Equinox',
                       guide = 'legend')+
  # Labels
  labs(title = 'Resultating segmentation')+
  xlab('Longitude') + ylab('Latitude')+
  # Correct legends colors
  guides(fill = guide_legend(override.aes = list(fill = breaks,
                                                 shape = 21)))+
  #Theme
  my_theme

breed_plot
```

### 4.4. Assign wintering areas

We project the track with the same projection that has the wintering areas shapefile. Per each wintering stage, we assign the ecoregion where more wintering points are located.

```{r}

# Track projected
track_sf <- track_f%>%
  st_as_sf(coords = c("Longitude", "Latitude"),
           crs = 4326, agr = "constant", remove = FALSE)

df_win <- st_join(track_sf, ecoregions, left = FALSE)

WAs <- df_win %>%
  filter(str_detect(Breed3, "^W\\d+$")) %>%  # Select wintering points
  group_by(Breed3) %>% 
  dplyr::summarise(
    Ecoregion = names(which.max(table(ECOREGION)))  # Which is the most frequent ecoregion
  ) %>%
  right_join(
    tibble(Breed3 = unique(df_win$Breed3[str_detect(df_win$Breed3, "^W\\d+$")])), 
    by = "Breed3"
  )

if(nrow(WAs) == 0){
  WAs <- ecoregions %>%
    filter(ECOREGION == 'Saharan_Upwelling')%>%
    add_column(Breed3 = 'W1')%>%
    dplyr::select(Breed3, Ecoregion = ECOREGION)
  
}
```

### 4.5. Detect resident individuals

Some species are partially migrant. What this means is, although the majority of the individuals migrate, part of them stay in areas close to their breeding area.
```{r}
if(all(track_f$Breed3 == 'B1')){
  Migrant <- 'Resident'
} else { 
  Migrant <- 'Migrant'
}
```

### 4.6. KDE phenology determination 

Before do the loop, we separate into a list each stationary stage of the track (breeding and wintering periods) using the function `phen_periods`. These periods include the stationary stage (B or W) joined to the previous and following migratory periods. What that means, to determine the wintering area we calculate the 90% KDE of mig1 + W1 + mig2, whereas for the pre-migration breeding area, we join B1 + mig1.

The function `calculate_kde` joins the process of the package `track2KBA`, utilizing "href" as smoothing factor and 90 as levelUD. Then, the funtion `find_dates` calculates the **first point inside as the arrival** point and **the first point outside the KDE as the departure** point. Knowing the points, the own function determines the dates of those points.

```{r}

ph_seg.list <- subset_to_kde(track_f, "Breed3")

# Loop for each stationary stage (breeding and wintering areas)
results_list <- list()
for (k in names(ph_seg.list)) {
  # k = 'B1'
  print(k)
  ph_seg <- ph_seg.list[[k]]
  
  rle_ph3 <- rle(df_format[['Breed3']])
  
  stg <- ph_seg[!(ph_seg$Breed3 %in% c("mig1", "mig2")), ]
  
  if (is.null(ph_seg) || nrow(stg) == 0) {
    # If no data, NA values
    results_list[[k]] <- list(KDE90_sf = NULL,
                              arr_point = data.frame(Longitude = NA, Latitude = NA), 
                              arr_date = NA,
                              dep_point = data.frame(Longitude = NA, Latitude = NA), 
                              dep_date = NA
    )
    next
  }
  
  # Calculate KDE for non-migratory periods
  KDE90_sf <- calculate_kde(ph_seg, proj = 4326)
  
  # Find first and last points inside KDE
  phen_dates <- find_dates(ph_seg, KDE90_sf, proj = 4326)
  
  # Store results
  results_list[[k]] <- c(list(KDE90_sf = KDE90_sf), phen_dates)
}
```

When we work with partial migratory species, the following lines are recommended to be sure that 
```{r}
if(Migrant == 'Resident'){
  results_list[["B1"]][["dep_date"]] <- as.Date(NA)
}
```

### 4.7. Plot phenology
The map summarize the results of the phenological dates extraction.
```{r}
map_ph <-  phen_map(track = track_f, col.lat = col.lat, col.lon = col.lon, results_list, world)

map_ph
```

### 4.8. Saving phenological dates
We store the data in a data frame that can be 
```{r}
phen_KDEdates <- extract_phen_dates(results_list)

phen <- data.frame(Ring = ring, Sp = sp, Colony = col.name,
                   ID = geo_trip, Migrant = Migrant, 
                   Year1 = start_year, Year2 = end_year,
                   phen_KDEdates
)
```

When we are working in loops we can save the results of phenology extraction in a list

```{r, eval=FALSE, include=TRUE}
phen.list[i] <-list(phen.df)
phen.all <- plyr::ldply(phen.list, data.frame)
```

```{r}
write_csv(phen.all, file = 'output/Phenology.csv')
```

